---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Split-plot-factorial ANOVA (SPF-p.q design)"
categories: [Univariate, ANOVA]
rerCat: Univariate
tags: [ANOVA]
---

Split-plot-factorial ANOVA (SPF-p.q design)
=========================

TODO
-------------------------

 - link to anovaSPFpqr, anovaMixed, dfReshape

Traditional univariate analysis and multivariate approach.

Install required packages
-------------------------

[`car`](https://cran.r-project.org/package=car), [`DescTools`](https://cran.r-project.org/package=DescTools), [`multcomp`](https://cran.r-project.org/package=multcomp)

```{r}
wants <- c("car", "DescTools", "multcomp")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Two-way SPF-\(p \cdot q\) ANOVA
-------------------------

### Using `aov()` with data in long format

```{r}
set.seed(123)
Nj   <- 10
P    <- 3
Q    <- 3
muJK <- c(rep(c(1,-1,-2), Nj), rep(c(2,1,-1), Nj), rep(c(3,3,0), Nj))
dfSPFpqL <- data.frame(id=factor(rep(1:(P*Nj), times=Q)),
                       IVbtw=factor(rep(LETTERS[1:P], times=Q*Nj)),
                       IVwth=factor(rep(1:Q, each=P*Nj)),
                       DV=rnorm(Nj*P*Q, muJK, 3))
```

```{r}
aovSPFpq <- aov(DV ~ IVbtw*IVwth + Error(id/IVwth), data=dfSPFpqL)
summary(aovSPFpq)
```

### Using `Anova()` from package `car` with data in wide format

```{r}
dfSPFpqW <- reshape(dfSPFpqL, v.names="DV", timevar="IVwth",
                    idvar=c("id", "IVbtw"), direction="wide")
```

```{r}
library(car)
fitSPFpq   <- lm(cbind(DV.1, DV.2, DV.3) ~ IVbtw, data=dfSPFpqW)
inSPFpq    <- data.frame(IVwth=gl(Q, 1))
AnovaSPFpq <- Anova(fitSPFpq, idata=inSPFpq, idesign=~IVwth)
summary(AnovaSPFpq, multivariate=FALSE, univariate=TRUE)
```

### Using `anova.mlm()` and `mauchly.test()` with data in wide format

```{r}
anova(fitSPFpq, M=~1, X=~0, idata=inSPFpq, test="Spherical")
anova(fitSPFpq, M=~IVwth, X=~1, idata=inSPFpq, test="Spherical")
```

```{r}
mauchly.test(fitSPFpq, M=~IVwth, X=~1, idata=inSPFpq)
```

Effect size estimates: generalized \(\hat{\eta}_{g}^{2}\)
-------------------------

```{r}
library(DescTools)
EtaSq(aovSPFpq, type=1)
```

Simple effects
-------------------------

Separate error terms

### Between-subjects effect at a fixed level of the within-subjects factor

```{r}
summary(aov(DV ~ IVbtw, data=dfSPFpqL, subset=(IVwth==1)))
summary(aov(DV ~ IVbtw, data=dfSPFpqL, subset=(IVwth==2)))
summary(aov(DV ~ IVbtw, data=dfSPFpqL, subset=(IVwth==3)))
```

### Within-subjects effect at a fixed level of the between-subjects factor

```{r}
summary(aov(DV ~ IVwth + Error(id/IVwth), data=dfSPFpqL,
        subset=(IVbtw=="A")))
summary(aov(DV ~ IVwth + Error(id/IVwth), data=dfSPFpqL,
        subset=(IVbtw=="B")))
summary(aov(DV ~ IVwth + Error(id/IVwth), data=dfSPFpqL,
        subset=(IVbtw=="C")))
```

Planned comparisons for the between-subjects factor
-------------------------

```{r}
mDf    <- aggregate(DV ~ id + IVbtw, data=dfSPFpqL, FUN=mean)
aovRes <- aov(DV ~ IVbtw, data=mDf)

```{r}
cMat <- rbind("-0.5*(A+B)+C"=c(-1/2, -1/2, 1),
                       "A-C"=c(-1,    0,   1))
```

```{r}
library(multcomp)
summary(glht(aovRes, linfct=mcp(IVbtw=cMat), alternative="greater"),
        test=adjusted("none"))
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:DescTools))
try(detach(package:multcomp))
try(detach(package:mvtnorm))
try(detach(package:car))
try(detach(package:survival))
try(detach(package:splines))
try(detach(package:TH.data))
```
