---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Basic arithmetic and logic"
categories: [RBasics]
rerCat: R_Basics
tags: [Arithmetic, Logic]
---

Basic arithmetic and logic
=========================

TODO
-------------------------

 - link to vector

R as a pocket calculator
-------------------------

### Arithmetic operators

```{r}
3 + 7
9 / 3
9 * (3+2)
12^2 + 1.5*10
10 %/% 3
10 %% 3
```

Also see `help(Syntax)` for operator precendence / associativity: This determines the order in which computations are carried out when multiple operators are present.

Using operators as functions

```{r}
"/"(1, 10)
"+"(2, 3)
```

### Standard math functions

```{r}
sqrt(4)
sin(pi/2)
abs(-4)
log10(100)
exp(1)
```

### Rounding numbers

```{r}
round(1.271)
round(pi, digits=3)
ceiling(1.2)
floor(3.7)
trunc(22.913)
```

### Complex numbers

```{r}
exp(1)^((0+1i)*pi)
exp(1)^(-pi/2) - (0+1i)^(0+1i)
sqrt(-1)
sqrt(-1+0i)
```

### Special values

```{r}
.Machine$integer.max
.Machine$double.eps
```

```{r}
1/0
is.infinite(1/0)
0/0
is.nan(0/0)
NULL
is.null(NULL)
```

### Using variables (objects)

```{r}
x1 <- 2
x2 <- 10
x3 <- -7
x1 * 2
x2^x1 + x3
```

Logic
-------------------------

### Logical values

```{r}
TRUE
FALSE
!TRUE
!FALSE
```

```{r}
isTRUE(TRUE)
isTRUE(FALSE)
```

### Logical comparisons

```{r}
TRUE == TRUE
TRUE == FALSE
TRUE != TRUE
TRUE != FALSE
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
FALSE & TRUE
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
FALSE | TRUE
xor(TRUE, FALSE)
xor(TRUE, TRUE)
```

Short-circuit logical comparisons with vectors

```{r}
c(TRUE,  FALSE, FALSE) && c(TRUE,  TRUE, FALSE)
c(FALSE, FALSE, TRUE)  || c(FALSE, TRUE, FALSE)
```

### Arithmetic comparisons

```{r}
4 < 8
7 < 3
4 > 4
4 >= 4
```

### Checking whether any or all elements are `TRUE`

```{r}
any(c(FALSE, FALSE, FALSE))
any(c(FALSE, FALSE, TRUE))
all(c(TRUE, TRUE, FALSE))
any(c(TRUE, TRUE, TRUE))
```

In an empty vector, there is no element that is `FALSE`, therefore:

```{r}
all(numeric(0))
```

In an empty vector, you cannot pick an element that is `TRUE`, therefore:

```{r}
any(numeric(0))
```

Numeric representations
-------------------------

### Integers vs. decimal numbers

```{r}
4L == 4
identical(4L, 4)
```

### Floating point arithmetic

```{r}
0.1 + 0.2 == 0.3
1 %/% 0.1
sin(pi)
1 - ((1/49) * 49)
1 - ((1/48) * 48)
```

[What every computer scientist should know about floating-point arithmetic](http://docs.sun.com/source/806-3568/ncg_goldberg.html)

### Checking decimal numbers for equality

```{r}
isTRUE(all.equal(0.123450001, 0.123450000))
0.123400001 == 0.123400000
all.equal(0.12345001, 0.12345000)
isTRUE(all.equal(0.12345001,  0.12345000))
```
