---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Data import and export"
categories: [DataFrames]
rerCat: Data_Frames
tags: [DataFrames]
---

Data import and export
=========================

Install required packages
-------------------------

[`haven`](https://cran.r-project.org/package=haven), [`readxl`](https://cran.r-project.org/package=readxl), [`writexl`](https://cran.r-project.org/package=writexl), [`RSQLite`](https://cran.r-project.org/package=RSQLite)

```{r}
wants <- c("haven", "readxl", "writexl", "RSQLite")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Read and write data in text format
-------------------------

### Read and write text data from a local file

If reading from a file, check with `getwd()` that you are in the correct directory - or specify full file path. If the result should be a data frame, use `read.table()`.

```{r eval=FALSE}
## write text file first
myDf <- data.frame(IV=factor(rep(c("A", "B"), 5)),
                   DV=rnorm(10))
write.table(myDf, file="data.txt", row.names=FALSE)

## read this text file
myDf_read <- read.table("data.txt", header=TRUE)
str(myDf_read)
## not shown
```

Some relevant options for `read.table()`

```{r eval=FALSE}
read.table(file="data.txt", header=TRUE)
read.table(file="data.txt", header=FALSE)
read.table(file="data.txt", sep="\t")
read.table(file="data.txt", na.strings="9999")
read.table(file="data.txt", stringsAsFactors=TRUE)
# not shown
```

Some relevant options for `write.table()`

```{r eval=FALSE}
write.table(file="data.txt", row.names=FALSE, col.names=TRUE)
write.table(file="data.txt", sep="\t")
# not shown
```

To read comma-separated-value files, use `read.csv()`, for fixed-with-files `read.fwf()`. To read a raw (unstructured) text file, use `readLines()`. The result is a character vector with one element per line.

### Read text data from other sources

The described R functions can also read data from a character string.

```{r}
txt <- 'X Y Z
123 "A B C" 34.8
99 "D E F" 3543.2'
read.table(text=txt, header=TRUE)
```

Other sources can be standard-input (if R is used in batch mode via `Rscript`), the clipboard, or directly from an online address.

```{r eval=FALSE}
read.table(file=stdin(), header=TRUE)
## Windows: file="clipboard", MacOS: file=pipe("pbpaste")
read.table(file="clipboard", header=TRUE, sep="\t"))
read.table(file="http://dwoll.de/r/data.txt", header=TRUE)
# not shown
```

Read and write data in R binary format
-------------------------

Save multiple objects together with their respective symbol (=name). Object names do not need to be assigned when reading data.

```{r eval=FALSE}
myDf <- data.frame(IV=factor(rep(c("A", "B"), 5)), DV=rnorm(10))
save(myDf, file="data.RData")
load("data.RData")
# not shown
```

Save objects separately without their symbol (=name) -> preferred. Object name needs to be assigned when reading data.

```{r eval=FALSE}
myDf <- data.frame(IV=factor(rep(c("A", "B"), 5)), DV=rnorm(10))
saveRDS(myDf, file="data.rda")
myDf_read <- readRDS("data.rda")
# not shown
```

Use R for data entry
-------------------------

```{r}
myVar <- c(4, 19, 22)
```

Read data from the console with `scan()`. Lines are terminated by hitting the `Return` key, `scan()` quits when `Return` is hit on a blank line.

```{r eval=FALSE}
vec     <- scan()
charVec <- scan(what=character())
# not shown
```

Save R data to file
-------------------------

### Create a logfile for commands and output

```{r eval=FALSE}
sink("d:/daniel/logfile.txt", split=TRUE)
# not shown
```

### Save and read R objects in text format

```{r eval=FALSE}
dump("myDf", file="dumpMyDf.txt")
source("dumpMyDf.txt")
# not shown
```

Exchange data with other statistics software and spreadsheets
-------------------------

### Exchange data with SPSS, SAS and Stata

One option is to use text files (tab-separated or comma-separated) as described above to exchange data with other statistics software packages.

Another option to exchange data with SPSS, SAS and Stata (among others) is the `haven` package. Example for SPSS:

```{r eval=FALSE}
library(haven)
write_sav(myDf, path="myDf.sav")
myDf_sav <- read_sav("myDf.sav")
head(myDf_sav, n=4)
str(myDf_sav)
## not shown
```

SPSS has the concept of "variable labels" and "value labels" which do not directly map to R concepts. Variables read from SPSS data files therefore have special labels and are of class `haven_labelled`. Categorical variables can be converted to regular R factors using `as_factor()`.

```{r eval=FALSE}
myDf_org <- transform(myDf, IV=as_factor(IV))
str(myDf_org)
## not shown
```

When writing data to an SPSS file, labels can be set manually, and the class can be changed to `haven_labelled`.

```{r eval=FALSE}
myDf <- data.frame(IV=factor(rep(c("A", "B"), 5)),
                   DV=rnorm(10),
                   intVar=sample(1:3, 10, replace=TRUE))
attr(myDf$IV, "label") <- "Grouping factor"
attr(myDf$DV, "label") <- "Measurement values"
class(myDf$intVar) <- "haven_labelled"
attr(myDf$intVar, "labels") <- c("Val1"=1, "Val2"=2, "Val3"=3)

library(haven)
write_sav(myDf, "myDf.sav")
```

### Use the SPSS Integration Plug-in for R

If you have SPSS available, install the "Integration plug-in for R" add-in. This allows you to run R within SPSS. The plug-in includes an R package with functions that transfer the active SPSS data frame to R (and back) - including labeled factor levels, dates and German umlauts. Once installed, you can use it like this in the SPSS syntax window:

```
BEGIN PROGRAM R.
# from here on, you can use R syntax
myDf <- spssdata.GetDataFromSPSS(missingValueToNA=TRUE,
                                 factorMode="labels",
                                 rDate="POSIXct")
save(myDf, file="d:/path/to/your/myDf.Rdata")
END PROGRAM.
```

### Exchange data with Excel

One option is to use text files (tab-separated or comma-separated) as described above to exchange data with spreadsheet applications.

```{r}
library(writexl)
write_xlsx(myDf, path="myDf.xlsx")

library(readxl)
myDf_read <- read_xlsx("myDf.xlsx")
```

Read and write data from a database
-------------------------

There are R packages that provide an interface to all common database types. Using databases is described in more detail in:

 * Adler, J. (2012). R in a Nutshell (2nd ed.). Sebastopol, CA: O'Reilly.
 * Spector, P. (2008). Data Manipulation with R. New York, NY: Springer.

SQL commands are described [here](https://www.1keydata.com/sql/) and [here](https://sqlzoo.net/).

### Using the DBI interface with `RSQLite`

Simulate data

```{r}
IQ     <- rnorm(2*10, mean=100, sd=15)
rating <- sample(LETTERS[1:3], 2*50, replace=TRUE)
sex    <- factor(rep(c("f", "m"), times=50))
myDf   <- data.frame(sex, IQ, rating, stringsAsFactors=FALSE)
```

Save data frame in SQLite database. This is usually a file. In this example, the file is created in memory only. Use `dbConnect(<driver object>, dbname="file_name.db")` to create a file on disk.

```{r}
library("RSQLite")
drv <- dbDriver("SQLite")
con <- dbConnect(drv, dbname=":memory:")
dbWriteTable(con, name="MyDataFrame", value=myDf, row.names=FALSE)
```

Find out which tables are present, and which fields are in a specific table.

```{r}
dbListTables(con)
dbListFields(con, "MyDataFrame")
```

Read complete table, then send SQL-query.

```{r}
out <- dbReadTable(con, "MyDataFrame")
head(out, n=4)
dbGetQuery(con, "SELECT sex, AVG(IQ) AS mIQ, SUM(IQ) AS sIQ FROM MyDataFrame GROUP BY sex")
```

Query database and read results in smaller partial chunks. Useful for large queries.

```{r}
res <- dbSendQuery(con, "SELECT IQ, rating FROM MyDataFrame WHERE rating = 'A'")

while(!dbHasCompleted(res)) {
  partial <- dbFetch(res, n=4)
  print(partial)
}
```

Clean query, remove the created table, and close the database connection.

```{r}
dbClearResult(res)
dbRemoveTable(con, "MyDataFrame")
dbDisconnect(con)
```

Useful documents
-------------------------

 * [R Data Import/Export](https://cran.at.r-project.org/doc/manuals/R-data.html)
 * Muenchen, R. A. (2011). R for SAS and SPSS Users (2nd ed.). New York, NY: Springer. [URL](http://r4stats.com/)

Useful packages
-------------------------

The [CRAN Web Technologies Task View](https://CRAN.R-project.org/view=WebTechnologies) presents packages to directly scrape data from online sources. [`data.table`](https://cran.r-project.org/package=data.table) provides function `fread()` for high performance reading of large plain text data files.

Detach (automatically) loaded packages (if possible)
-------------------------

```{r eval=FALSE}
try(detach(package:haven))
try(detach(package:readxl))
try(detach(package:writexl))
try(detach(package:RSQLite))
try(detach(package:DBI))
```
