---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Choose names, variables, cases - dplyr"
categories: [DataFrames]
rerCat: Data_Frames
tags: [DataFrames]
---

Choose names, variables, cases of data frames - package `dplyr`
=========================

Install required packages
-------------------------

[`dplyr`](https://cran.r-project.org/package=dplyr), [`forcats`](https://cran.r-project.org/package=forcats)

```{r}
wants <- c("dplyr", "forcats")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

TODO
-------------------------

 - link to strings for `grep()`, dfSplitMerge, dfReshape

Simulate data
-------------------------

```{r}
set.seed(123)
N      <- 12
sex    <- factor(sample(c("f", "m"), N, replace=TRUE), levels=c("f", "m"))
group  <- factor(sample(rep(c("CG", "WL", "T"), 4), N, replace=FALSE), levels=c("CG", "WL", "T"))
age    <- sample(18:35, N, replace=TRUE)
IQ     <- round(rnorm(N, mean=100, sd=15))
rating <- round(runif(N, min=0, max=6))
(myDf1 <- data.frame(id=1:N, sex, group, age, IQ, rating))
```

Rename variables
-------------------------

### Specify new and old names explicitly

```{r}
library(dplyr)
myDf1 %>%
    rename(score=rating,
           fac=group)
```

### Specify new and old name programmatically

Make all names lower case that start with 'i' or 'I'

```{r}
myDf1 %>%
    rename_with(tolower, matches("^i.+"))
```

Prefix all `character` variables with `fac_`

```{r}
myDf1 %>%
    rename_with(~paste0("fac_", .), where(is.character))
```

Make all variable names upper case

```{r}
myDf1 %>%
    rename_with(toupper)
```

Select variables
-------------------------

### Select by full name

```{r}    
library(dplyr)
myDf1 %>%
    select(group, IQ)
```

### Select by name pattern

```{r}
myDf1 %>%
    select(matches("^i.+"))
```

### Select by data type

```{r}
myDf1 %>%
    select(where(is.numeric))
```

### Select by logical combination of characteristics

```{r}
myDf1 %>%
    select(!where(is.numeric) & starts_with("g"))
```

### De-select variables

```{r}
myDf1 %>%
    select(-sex, -matches("a.+"))

myDf1 %>%
    select(-(group:IQ))
```

Select cases
-------------------------

### By explicitly stating matching criteria

Matching all criteria simultaneously - logical AND

```{r}    
myDf1 %>%
    filter(group %in% c("CG", "WL"),
           id > rating,
           IQ > 90)
```

Matching some criteria - logical OR

```{r}
myDf1 %>%
    filter((IQ < 90) | (IQ > 110))
```

By row number

```{r}
myDf1 %>%
    slice(5:7)
```

Drop duplicate rows

```{r}
myDf1 %>%
    select(sex, group) %>%
    distinct()
```

```{r}
myDf1 %>%
    distinct(sex, group)
```

Count distinct rows

```{r}
myDf1 %>%
    n_distinct()
```

### Programmatically stating matching criteria

Add some missing values to data frame first

```{r}
getNA <- function(x, prob=c(0.7, 0.3)) {
    NAval <- x[length(x) + 1]
    if_else(sample(c(TRUE, FALSE), length(x), replace=TRUE, prob=prob), x, NAval)
}

set.seed(123)
myDf1NA <- myDf1 %>%
    mutate(group=getNA(group),
           age=getNA(age),
           IQ=getNA(IQ),
           rating=getNA(rating))

myDf1NA
```

Show only cases with all missings in variables "from group to IQ"

```{r}
myDf1NA %>%
    filter(across(group:IQ, is.na))
```

Show only cases with uneven values on all numeric variables

```{r}
myDf1NA %>%
    filter(across(where(is.numeric), ~ (. %% 2) == 1))
```

Change data frame variables
-------------------------

### Explicitly

Add new variables or overwrite existing variables in `mutate()`

```{r}
library(dplyr)
library(forcats)

myDf1 %>%
    mutate(married=sample(c(TRUE, FALSE), n(), replace=TRUE))

myDf1 %>%
    mutate(group=fct_collapse(group, CG_WL=c("CG", "WL")),
           ratingSq=rating^2,
           ratingSqZ=scale(ratingSq))
```

Conditional changes to variables

```{r}
myDf1 %>%
    mutate(age_even=if_else((age %% 2) == 0, TRUE, FALSE),
           sex_IQ=case_when(
               ((sex == "f") & (IQ <  100)) ~ "female_lo",
               ((sex == "f") & (IQ >= 100)) ~ "female_hi",
               ((sex == "m") & (IQ <  100)) ~ "male_lo",
               ((sex == "m") & (IQ >= 100)) ~ "male_hi",
               TRUE                         ~ "other"))
```

Recode special values to missing (`NA`)

```{r}
myDf9999           <- myDf1
myDf9999$IQ[2]     <- 9999
myDf9999$rating[3] <- 9999

(myDfNA <- myDf9999 %>%
    mutate(IQ=na_if(IQ, 9999),
           rating=na_if(rating, 9999)))
```

Drop missing values

```{r}
myDfNA %>%
    na.omit()
```

### Programmatically

```{r}
myDf1 %>%
    mutate(across(where(is.numeric),
                  ~scale(., center=TRUE, scale=FALSE)))

myDf1 %>%
    mutate(across(age:IQ, list(ctr=~scale(., center=TRUE, scale=FALSE),
                               scl=~scale(., center=FALSE, scale=TRUE))))

myDf1 %>%
    mutate(across(everything(), getNA))
```


Change order of rows
-------------------------

Ascending

```{r}
myDf1 %>%
    arrange(rating)
```

Descending

```{r}
myDf1 %>%
    arrange(group, desc(IQ))
```

Change order of variables
-------------------------

Move to front

```{r}
myDf1 %>%
    select(group, age, everything())
```

Specify position directly

```{r}
myDf1 %>%
    relocate(group, .after=rating)
```


Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:dplyr))
try(detach(package:forcats))
```
