---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Choose names, variables, cases - base R"
categories: [DataFrames]
rerCat: Data_Frames
tags: [DataFrames]
---

Choose names, variables, cases - base R
=========================

TODO
-------------------------

 - link to strings for `grep()`, dfSplitMerge, dfReshape

Simulate data
-------------------------

```{r}
set.seed(123)
N      <- 12
sex    <- factor(sample(c("f", "m"), N, replace=TRUE), levels=c("f", "m"))
group  <- factor(sample(rep(c("CG", "WL", "T"), 4), N, replace=FALSE), levels=c("CG", "WL", "T"))
age    <- sample(18:35, N, replace=TRUE)
IQ     <- round(rnorm(N, mean=100, sd=15))
rating <- round(runif(N, min=0, max=6))
(myDf1 <- data.frame(id=1:N, sex, group, age, IQ, rating))
```

Rename variables
-------------------------

```{r}
names(myDf1)
names(myDf1)[3] <- "fac"
names(myDf1)
names(myDf1)[names(myDf1) == "fac"] <- "group"
names(myDf1)
```

Select variables
-------------------------

### Select by full name

```{r}
subset(myDf1, select=c(group, IQ))
```

### Select by name pattern

```{r}
(colIdx <- grep("^i.+", names(myDf1), ignore.case=TRUE))
subset(myDf1, select=colIdx)
```

### Select by data type

Only numeric variables

```{r}
(colIdx <- sapply(myDf1, is.numeric))
subset(myDf1, select=colIdx)
```

More elegant

```{r}
myDfNum <- Filter(is.numeric, myDf1)
head(myDfNum)
```

### De-select variables

```{r}
subset(myDf1, select=c(-sex, -IQ))
```

Remove variables - 1 variable

```{r}
dfTemp <- myDf1
dfTemp$group <- NULL
head(dfTemp, n=3)
```

Remove variables - multiple variables

```{r}
dfTemp[c("sex", "IQ")] <- list(NULL)
head(dfTemp, n=3)
```

Select cases
-------------------------

### Select by logical conditions

Simple conditions

```{r}
subset(myDf1, sex == "f")
subset(myDf1, id == rating)
```

Combined conditions - logical AND

```{r}
subset(myDf1, (sex == "m") & (rating > 2))
```

Combined conditions - logical OR

```{r}
subset(myDf1, (IQ < 90) | (IQ > 110))
subset(myDf1, group %in% c("CG", "WL"))
```

### Identify and drop duplicates

```{r}
myDfDouble <- rbind(myDf1, myDf1[sample(seq_len(nrow(myDf1)), 4), ])
duplicated(myDfDouble) | duplicated(myDfDouble, fromLast=TRUE)
myDfUnique <- unique(myDfDouble)
any(duplicated(myDfUnique))
```

### Identify, count, and drop cases with missings


```{r}
myDfNA           <- myDf1
myDfNA$IQ[2]     <- NA
myDfNA$rating[3] <- NA
```

Count missings

```{r}
is.na(myDfNA)[1:3, ]
apply(myDfNA, 2, anyNA)
```

Count missings / present observations by variable

```{r}
apply(myDfNA, 2, function(x) { sum(is.na(x)) })
apply(myDfNA, 2, function(x) { length(na.omit(x)) })
```

Drop cases with missings

```{r}
complete.cases(myDfNA)
head(na.omit(myDfNA), n=4)
table(complete.cases(myDfNA))
```

Show only cases with missings

```{r}
subset(myDfNA, !complete.cases(myDfNA))
```

Change data frame variables
-------------------------

### Add new variables or overwrite existing variables

```{r}
married <- sample(c(TRUE, FALSE), nrow(myDf1), replace=TRUE)
myDf2   <- myDf1
myDf2$married1    <- married
myDf2["married2"] <- married
head(myDf2)
```

```{r}
myDf3 <- cbind(myDf1, married)
head(myDf3, n=3)
```

```{r}
myDf4 <- transform(myDf3,
                   rSq=rating^2,
                   IQgrp=cut(IQ, breaks=c(0, 100, Inf)))
head(myDf4, n=3)
```

Change order of rows
-------------------------

```{r}
(idx1 <- order(myDf1$rating))
myDf1[idx1, ]
(idx2 <- order(myDf1$group, myDf1$IQ))
myDf1[idx2, ]
```

Useful packages
-------------------------

In order to work with data frames like with a database, use [`sqldf`](https://cran.r-project.org/package=sqldf). You can then use standard SQL commands to select data. [`data.table`](https://cran.r-project.org/package=data.table) provides an alternative (compatible) class of data frames with better performance for large amounts of data.
