---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Diagram types with the ggplot2 package"
categories: [Diagrams, BasicDiagrams]
rerCat: Diagrams
tags: [Diagrams, ggplot2]
---

Diagrams with the `ggplot2` package
=========================

Install required packages
-------------------------

[`ggplot2`](https://cran.r-project.org/package=ggplot2), [`ggbeeswarm`](https://cran.r-project.org/package=ggbeeswarm), [`cowplot`](https://cran.r-project.org/package=cowplot), [`patchwork`](https://cran.r-project.org/package=patchwork)

```{r}
wants <- c("ggplot2", "ggbeeswarm", "cowplot", "patchwork")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Simulate data
-------------------------

```{r}
Njk    <- 50
P      <- 4
Q      <- 2
IQ     <- rnorm(P*Q*Njk, mean=100, sd=15)
height <- rnorm(P*Q*Njk, mean=175, sd=7)
rating <- factor(sample(LETTERS[1:3], Njk*P*Q, replace=TRUE))
sex    <- factor(rep(c("f", "m"), times=P*Njk))
group  <- factor(rep(c("control", "placebo", "treatA", "treatB"), each=Q*Njk))
sgComb <- interaction(sex, group)
mood   <- round(rnorm(P*Q*Njk, mean=c(85, 80, 110, 90, 130, 100, 93, 121)[sgComb], sd=25))
myDf   <- data.frame(sex, group, sgComb, IQ, height, rating, mood)
```

Diagram types
-------------------------

### Scatterplot

```{r  ggplot_types01}
library(ggplot2)
ggplot(myDf, aes(x=height, y=IQ)) +
        geom_point(size=3)
```

### Line plot

`group=TRUE` is necessary here

```{r ggplot_types02a}
(groupM <- aggregate(mood ~ group, data=myDf, FUN=mean))
ggplot(groupM, aes(x=group, y=mood, group=TRUE)) +
    geom_point(size=3) +
    geom_line()
```

```{r ggplot_types02b}
(groupM <- aggregate(mood ~ group, data=myDf, FUN=mean))
ggplot(groupM, aes(x=group, y=mood, group=TRUE)) +
    geom_point(size=3) +
    geom_step()
```

### Barplot

Absolute frequencies

```{r  ggplot_types03}
ggplot(myDf, aes(x=rating)) +
    geom_bar(stat="count") +
        ggtitle("Rating frequencies")
```

Relative frequencies

```{r ggplot_types04}
ggplot(myDf, aes(x=rating)) +
    geom_bar(stat="count",
             aes(y=(..count..) / sum(..count..))) +
        ggtitle("Relative rating frequencies")
```

### Histogram

Absolute frequencies

```{r  ggplot_types06}
ggplot(myDf, aes(x=IQ)) +
        geom_histogram(bins=20) +
        ggtitle("Histogram IQ by group") +
        theme(legend.position="none")
```

Density with nonparametric kernel density estimator

```{r ggplot_types07}
ggplot(myDf, aes(x=mood)) +
    geom_histogram(aes(y=..density..)) +
    ggtitle("Histogram mood by group") +
    geom_density(color="darkgrey", fill="grey", alpha=0.6)
```

### Boxplot

Add a `geom_beeswarm()` layer to show raw data.

```{r ggplot_types08}
library(ggbeeswarm)
ggplot(myDf, aes(x=sex, y=height)) +
    geom_boxplot(outlier.shape=NULL) +
    geom_beeswarm(alpha=0.5) +
    ggtitle("Height by sex") +
    theme(legend.position="none")
```

### Q-Q plot

Against normal distribution

```{r ggplot_types09}
ggplot(myDf, aes(sample=height)) +
    geom_qq() +
    geom_qq_line(color="blue")
```

Against pre-specified distribution (here: t)

```{r ggplot_types10}
parL <- list(df=10, ncp=0)
ggplot(myDf, aes(sample=height)) +
    geom_qq(distribution=qt, dparams=parL) +
    geom_qq_line(color="blue", distribution=qt, dparams=parL)
```

Compare quantiles of two variables

```{r ggplot_types11}
probs <- ppoints(50)
qDat  <- data.frame(qmood=quantile(myDf$mood, probs),
                    qheight=quantile(myDf$height, probs))

ggplot(qDat, aes(x=qmood, y=qheight)) +
    geom_point() +
    xlab("Quantiles mood") +
    ylab("Quantiles height")
```

Multi-panel plots
-------------------------

Split the diagram into panels by column using `facet_grid()`.

```{r ggplot_types12a}
ggplot(myDf, aes(x=rating, y=IQ)) +
        geom_point(size=3) +
        facet_grid(. ~ group) +
        ggtitle("IQ ~ height split by sex and group") +
        guides(shape="none")
```

Split the diagram into matrix of panels using `facet_wrap()`. Option `scales="free_y"` produces separate y-axis for each panel.

```{r ggplot_types12b}
ggplot(myDf, aes(x=rating, y=IQ)) +
        geom_point(size=3) +
        facet_wrap(. ~ group, scales="free_y") +
        ggtitle("IQ ~ height split by sex and group") +
        guides(shape="none")
```

Additional diagram elements
-------------------------

Add error bars

```{r ggplot_types13}
groupN     <- as.data.frame(xtabs(~ sex + group, data=myDf))
groupM     <- aggregate(mood ~ sex + group, data=myDf, FUN=mean)
groupSD    <- aggregate(mood ~ sex + group, data=myDf, FUN=sd)
groupMSD   <- merge(groupM, groupSD, by=c("sex", "group"), suffixes=c(".M", ".SD"))
groupMSDN  <- merge(groupMSD, groupN, by=c("sex", "group"))
(groupMSDN <- transform(groupMSDN,
                        SEMlo=mood.M - mood.SD/sqrt(Freq),
                        SEMup=mood.M + mood.SD/sqrt(Freq)))

ggplot(groupMSDN, aes(x=group, y=mood.M, ymin=SEMlo, ymax=SEMup,
                              color=sex, shape=sex, group=sex)) +
    geom_point() +
    geom_line() +
    geom_linerange()
```

 * `geom_hline()` for a horizontal line
 * `geom_vline()` for a vertical line
 * `geom_smooth()` for a regression line including SEs
 * `geom_text()` for text from a variable
 * `ggtitle()` for a diagram title
 * `annotate()` for text given directly as an argument

```{r  ggplot_types14}
ggplot(myDf, aes(x=height, y=IQ)) +
        geom_hline(aes(yintercept=100), linetype=2) +
        geom_vline(aes(xintercept=180), linetype=2) +
        geom_point(size=3) +
        geom_smooth(method=lm, se=TRUE, size=1.2, fullrange=TRUE) +
        facet_grid(sex ~ group) +
        ggtitle("IQ ~ height split by sex and group") +
        theme(legend.position="none") +
        geom_text(aes(x=190, y=70, label=sgComb)) +
        annotate("text", x=165, y=130, label="annotation")
```

Put different plots on the same page
-------------------------

Using package `cowplot`

```{r ggplot_types15}
library(cowplot)
p1 <- ggplot(myDf, aes(x=height, y=IQ)) +
    geom_point()

p2 <- ggplot(myDf, aes(x=rating)) +
    geom_bar(stat="count", position=position_stack()) +
    ggtitle("Rating frequencies by sex and group")

plot_grid(p1, p2, ncol=1, align="v")
```

Using package `patchwork`

```{r ggplot_types16}
library(patchwork)
p1 + p2 + plot_layout(ncol=2)
```

```{r ggplot_types17}
p1 + p2 + plot_layout(ncol=1)
```

Save `ggplot2` diagrams to file
-------------------------

Graphics file format is determined from file ending.

```{r eval=FALSE}
p <- ggplot(myDf, aes(x=height, y=IQ)) +
        geom_point(size=3)

ggsave(p, file="diag_ggplot.pdf", width=9, height=6, units="in")
ggsave(p, file="diag_ggplot.png", width=200, height=150, units="mm", dpi=300)
```

Further online resources
-------------------------

 * See [Cookbook for R: `ggplot2` diagrams](https://r-graphics.org/) for many detailed examples of `ggplot2` diagrams.
 * Package [`esquisse`](https://cran.r-project.org/package=esquisse) provides an interactive Shiny app to create and customize ggplot2 diagrams.

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:cowplot))
try(detach(package:patchwork))
try(detach(package:ggbeeswarm))
try(detach(package:ggplot2))
```
