---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Format ggplot2 diagrams"
categories: [Diagrams]
rerCat: Diagrams
tags: [Diagrams, ggplot2]
---

Format diagrams with the `ggplot2` package
=========================

Install required packages
-------------------------

[`ggplot2`](https://cran.r-project.org/package=ggplot2), [`colorspace`](https://cran.r-project.org/package=colorspace)

```{r}
wants <- c("ggplot2", "colorspace")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Simulate data
-------------------------

Data needs to be in long format.

```{r}
Njk    <- 50
P      <- 3
Q      <- 2
IQ     <- rnorm(P*Q*Njk, mean=100, sd=15)
height <- rnorm(P*Q*Njk, mean=175, sd=7)
rating <- factor(sample(LETTERS[1:3], Njk*P*Q, replace=TRUE))
sex    <- factor(rep(c("f", "m"), times=P*Njk))
group  <- factor(rep(c("control", "placebo", "treatment"), each=Q*Njk))
sgComb <- interaction(sex, group)
mood   <- round(rnorm(P*Q*Njk, mean=c(85, 80, 110, 90, 130, 100)[sgComb], sd=25))
myDf   <- data.frame(sex, group, sgComb, IQ, height, rating, mood)
```

Control element position
-------------------------

Side-by-side = "dodge"

```{r ggplot_format01a}
library(ggplot2)
ggplot(myDf, aes(x=mood, fill=sex)) +
    geom_histogram(aes(y=..density..), position=position_dodge())
```

Random jitter - useful for overlapping points, e.g., when variables are integer.

```{r ggplot_format01b}
ggplot(myDf, aes(x=group, y=mood)) +
    geom_point()
```

```{r ggplot_format01c}
ggplot(myDf, aes(x=group, y=mood)) +
    geom_point(position=position_jitter(width=0.2))
```

```{r ggplot_format01d}
ggplot(myDf, aes(x=rating, group=sex, fill=sex)) +
    geom_bar(stat="count",
             aes(y=(..count..) / sum(..count..)),
             position=position_fill())
```

Format axes
-------------------------

Change order of categories by changing order of factor levels.

```{r ggplot_format02}
levels(myDf$rating)
levels(myDf$rating) <- rev(levels(myDf$rating))
ggplot(myDf, aes(x=rating, group=sex, fill=sex)) +
    geom_bar(stat="count",
             aes(y=(..count..) / sum(..count..)),
             position=position_fill()) +
    labs(x="Rating category", y="Cumulative relative frequency")
```

Rotate x-axis labels and fine tune x-axis limits / breaks.

```{r ggplot_format03}
ggplot(myDf, aes(x=height, y=mood, colour=sex, shape=group)) +
    geom_point(size=3) +
    scale_x_continuous(limits=c(150, 200),
                       expand=c(0, 0),
                       breaks=seq(150, 200, by=5)) +
    scale_y_continuous(n.breaks=8) +
    guides(x=guide_axis(angle=90))
```

Flip x- and y-axis

```{r ggplot_format04}
ggplot(myDf, aes(x=sex, y=height, fill=sex)) +
    geom_boxplot() +
    coord_flip(ylim=c(140, 200))
```

Format legend
-------------------------

```{r ggplot_format05}
ggplot(myDf, aes(x=height, y=mood, colour=sex:group, shape=sex)) +
    geom_hline(aes(yintercept=100), linetype=2) +
    geom_vline(aes(xintercept=180), linetype=2) +
    geom_point(size=3) +
    geom_smooth(method=lm, se=TRUE, size=1.2, fullrange=TRUE) +
    facet_grid(sex ~ group) +
    labs(title="mood ~ height getrennt nach Geschlecht + Gruppe") +
    geom_text(aes(x=190, y=70, label=sgComb), size=7, hjust="right") +
    annotate("text", x=165, y=35, size=7, label="Annotation") +
    guides(shape=FALSE) +
    theme(legend.position="bottom")
```

Format color, shape and line type
-------------------------

```{r ggplot_format06}
library(colorspace)
(groupM <- aggregate(mood ~ sex + group, data=myDf, FUN=mean))

ggplot(groupM, aes(x=group, y=mood, color=sex, shape=sex, group=sex)) +
    geom_point(size=8, stroke=2) +
    geom_line(size=2, linetype="dashed") +
    scale_shape_discrete(solid=FALSE)

ggplot(myDf, aes(x=mood, fill=group)) +
    geom_histogram(aes(y=..density..), alpha=0.5) +
    geom_density(alpha=0.7) +
    scale_fill_discrete_qualitative()
```

Use themes for a collection of pre-set formatting choices.
-------------------------

```{r ggplot_format07}
ggplot(myDf, aes(x=mood, fill=group)) +
    geom_histogram(aes(y=..density..), alpha=0.5) +
    geom_density(alpha=0.7) +
    scale_fill_discrete_qualitative() +
    theme_bw()
```

```{r ggplot_format08}
ggplot(myDf, aes(x=sex, y=height, fill=sex)) +
    geom_boxplot() +
    coord_flip(ylim=c(140, 200)) +
    theme_minimal()
```

Further resources
-------------------------

See [Cookbook for R: `ggplot2` diagrams](https://r-graphics.org/) for many detailed examples of `ggplot2` diagrams.

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:ggplot2))
try(detach(package:colorspace))
```
