---
title: "Convex hull, (minimum) bounding box, minimum enclosing ellipse, and minimum enclosing circle"
author: "Daniel Wollschlaeger"
license: Creative Commons BY-SA
categories:
- Diagrams
- SpecificDiagrams
rerCat: Diagrams
tags: [Diagrams]
---

Convex hull, (minimum) bounding box, minimum enclosing ellipse, and minimum enclosing circle
=========================

Convex hull
-------------------------

`chull()` returns the ordered indices of the matrix rows corresponding to corners of the convex hull.

```{r}
set.seed(123)
xy      <- matrix(rnorm(24, 100, 15), ncol=2)
hullIdx <- chull(xy)
hullIdx
```

```{r rerDiagBounding01}
plot(xy, xlab="x", ylab="y", asp=1, type="n")
polygon(xy[hullIdx, ], border="blue", lwd=2)
points(xy, pch=16, cex=1.5)
```

Bounding box
-------------------------

```{r}
getBoundingBox <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) >= 2, ncol(xy) == 2)
    x   <- range(xy[ , 1])
    y   <- range(xy[ , 2])
    pts <- c(xleft=x[1], ybottom=y[1], xright=x[2], ytop=y[2])
    return(list(pts=pts, width=abs(diff(x)), height=abs(diff(y))))
}
```

```{r rerDiagBounding02}
bb <- getBoundingBox(xy)
plot(xy, xlab="x", ylab="y", asp=1, type="n")
rect(bb$pts[1], bb$pts[2], bb$pts[3], bb$pts[4], border="blue", lwd="2")
points(xy, pch=16, cex=1.5)
```

```{r}
bb$width * bb$height
```

Minimum bounding box
-------------------------

### Rotating calipers algorithm

```{r}
getMinBBox <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) >= 2, ncol(xy) == 2)

    ## rotating calipers algorithm using the convex hull
    H    <- chull(xy)                    # hull indices, vertices ordered clockwise
    n    <- length(H)                    # number of hull vertices
    hull <- xy[H, ]                      # hull vertices

    ## unit basis vectors for all subspaces spanned by the hull edges
    hDir  <- diff(rbind(hull, hull[1,])) # account for circular hull vertices
    hLens <- sqrt(rowSums(hDir^2))       # length of basis vectors
    huDir <- diag(1/hLens) %*% hDir      # scaled to unit length

    ## unit basis vectors for the orthogonal subspaces
    ## rotation by 90 deg -> y' = x, x' = -y
    ouDir <- cbind(-huDir[ , 2], huDir[ , 1])

    ## project hull vertices on the subspaces spanned by the hull edges, and on
    ## the subspaces spanned by their orthogonal complements - in subspace coords
    projMat <- rbind(huDir, ouDir) %*% t(hull)

    ## range of projections and corresponding width/height of bounding rectangle
    rangeH  <- matrix(numeric(n*2), ncol=2)   # hull edge
    rangeO  <- matrix(numeric(n*2), ncol=2)   # orth subspace
    widths  <- numeric(n)
    heights <- numeric(n)
    for(i in seq(along=H)) {
        rangeH[i, ] <- range(projMat[  i, ])
        rangeO[i, ] <- range(projMat[n+i, ])  # orth subspace is in 2nd half
        widths[i]   <- abs(diff(rangeH[i, ]))
        heights[i]  <- abs(diff(rangeO[i, ]))
    }

    ## extreme projections for min-area rect in subspace coordinates
    eMin  <- which.min(widths*heights)   # hull edge leading to minimum-area
    hProj <- rbind(   rangeH[eMin, ], 0)
    oProj <- rbind(0, rangeO[eMin, ])

    ## move projections to rectangle corners
    hPts <- sweep(hProj, 1, oProj[ , 1], "+")
    oPts <- sweep(hProj, 1, oProj[ , 2], "+")

    ## corners in standard coordinates, rows = x,y, columns = corners
    ## in combined (4x2)-matrix: reverse point order to be usable in polygon()
    basis <- cbind(huDir[eMin, ], ouDir[eMin, ])  # basis formed by hull edge and orth
    hCorn <- basis %*% hPts
    oCorn <- basis %*% oPts
    pts   <- t(cbind(hCorn, oCorn[ , c(2, 1)]))

    return(list(pts=pts, width=widths[eMin], height=heights[eMin]))
}
```

### Draw the minimum bounding box

```{r rerDiagBounding03}
mbb <- getMinBBox(xy)       ## minimum bounding box
H   <- chull(xy)            ## convex hull

# plot original points, convex hull, and minimum bounding box
plot(xy, xlab="x", ylab="y", asp=1, type="n",
         xlim=range(c(xy[ , 1], mbb$pts[ , 1])),
         ylim=range(c(xy[ , 2], mbb$pts[ , 2])))
polygon(xy[H, ], col=NA)    ## show convex hull
polygon(mbb$pts, border="blue", lwd=2)
points(xy, pch=16, cex=1.5)
```

```{r}
mbb$width * mbb$height      ## box area
```

Minimum enclosing ellipse
-------------------------

### Kachiyan's algorithm

Adapted from Matlab code by [Jacob on Stack Overflow](https://stackoverflow.com/a/1768440).

```{r rerDiagBoundingEll1}
getMinEllipse <- function(xy, tol=0.001, max_iter=1000) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) >= 2, ncol(xy) == 2)

    H    <- chull(xy)   # convex hull indices (vertices ordered clockwise)
    hPts <- t(xy[H, ])  # points that make up the convex hull
    p    <- nrow(hPts)  # dimension
    N    <- ncol(hPts)  # number of points on hull
    Q    <- rbind(hPts, rep(1, N))
    iter <- 1           # iteration
    err  <- 1           # remaining error
    u    <- rep(1/N, N)

    # Khachiyan's algorithm
    while((err > tol) && (iter < max_iter)) {
        X         <- Q %*% diag(u) %*% t(Q)
        m         <- diag(t(Q) %*% solve(X) %*% Q)
        maximum   <- max(m)
        j         <- which.max(m)
        step_size <- (maximum - p-1) / ((p+1)*(maximum-1))
        new_u     <- (1 - step_size)*u
        new_u[j]  <- new_u[j] + step_size
        err       <- sqrt(sum((new_u-u)^2))
        iter      <- iter + 1
        u         <- new_u
    }
    
    if(iter >= max_iter) {
        warning(paste("Maximum number of iterations reached. Error is still:", err))
    }
    
    ctr   <- hPts %*% u                # ellipse center
    E     <- (1/p) * solve((hPts %*% diag(u) %*% t(hPts)) - tcrossprod(ctr))
    shape <- solve(E)
    area0 <- pi                        # area unit circle
    area  <- area0 * sqrt(det(shape))  # area ellipse
    rad   <- sqrt(eigen(shape)$values) # length ellipse semi axes

    list(ctr=ctr, shape=shape, area=area, rad=rad)
}
```

### Draw the minimum enclosing ellipse

```{r rerDiagBoundingEll2}
me     <- getMinEllipse(xy)
CF     <- chol(me$shape, pivot=TRUE)      # Cholesky-factor
CFord  <- order(attr(CF, "pivot"))
angles <- seq(0, 2*pi, length.out=100)    # angles in radians
ell    <- 1 * cbind(cos(angles), sin(angles)) %*% CF[ , CFord]  # ellipse
ellCtr <- sweep(ell, 2, me$ctr, "+")      # move ellipse to center

# determine axis limits so that the ellipse will be visible
xLims <- me$ctr[1] + c(-me$rad[1], me$rad[1])
yLims <- me$ctr[2] + c(-me$rad[1], me$rad[1])

# draw points and ellipse
plot(xy, xlab="x", ylab="y", xlim=xLims, ylim=yLims, asp=1, type="n")
points(xy, pch=16)
points(me$ctr[1], me$ctr[2], pch=4, cex=1.5, lwd=2, col="blue")  # center
polygon(ellCtr, border="blue")         # ellipse
```

Minimum enclosing circle
-------------------------

Skyum algorithm based on the convex hull

### Identify collinear points on convex hull

```{r}
## assumes that points in xy are already oredered
idCollinear <- function(xy) {
    if(!is.matrix(xy))  { stop("xy must be a matrix") }
    if(!is.numeric(xy)) { stop("xy must be numeric") }
    if(nrow(xy) < 3L)   { stop("xy must have at least 3 points") }
    if(ncol(xy) != 2L)  { stop("xy must have 2 columns") }

    n    <- nrow(xy)
    idx  <- seq_len(n)
    post <- (idx %% n) + 1              # next point in S
    prev <- idx[order(post)]            # previous point in S

    del <- integer(0)
    ## check all sets of 3 consecutive points if they lie in 1D sub-space
    for(i in idx) {
        pts <- rbind(xy[prev[i], ],
                     xy[i, ],
                     xy[post[i], ])
        pts_rank <- qr(scale(pts, center=TRUE, scale=FALSE))$rank
        if(pts_rank < 2L) {
            del[length(del) + 1] <- i
        }
    }

    del
}
```

### Circle defined by three points

```{r}
getCircleFrom3 <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) == 3, ncol(xy) == 2)

    aa <- xy[1,  ]
    bb <- xy[2,  ]
    cc <- xy[3,  ]
    y  <- xy[ , 2]

    xDeltaA <- bb[1] - aa[1]
    yDeltaA <- bb[2] - aa[2]
    xDeltaB <- cc[1] - bb[1]
    yDeltaB <- cc[2] - bb[2]
    xDeltaC <- cc[1] - aa[1]
    yDeltaC <- cc[2] - aa[2]

    ## check if the points are collinear: qr(xy)$rank == 1, or:
    ## determinant of difference matrix = 0, no need to use det()
    dMat <- rbind(c(xDeltaA, yDeltaA), c(xDeltaB, yDeltaB))
    if(isTRUE(all.equal(dMat[1,1]*dMat[2,2] - dMat[1,2]*dMat[2,1], 0, check.attributes=FALSE))) {
        ## define the circle as the one that's centered between the points
        rangeX <- range(c(aa[1], bb[1], cc[1]))
        rangeY <- range(c(aa[2], bb[2], cc[2]))
        ctr    <- c(rangeX[1] + 0.5*diff(rangeX), rangeY[1] + 0.5*diff(rangeY))
        rad    <- sqrt((0.5*diff(rangeX))^2 + (0.5*diff(rangeY))^2)
    } else {
        rad <- prod(dist(xy)) / (2 * abs(det(cbind(xy, 1))))  # circle radius
        v1  <- rowSums(xy^2)                    # first vector in the numerator
        v2x <- c( xDeltaB, -xDeltaC,  xDeltaA)  # 2nd vector numerator for Mx
        v2y <- c(-yDeltaB,  yDeltaC, -yDeltaA)  # 2nd vector numerator for My
        ctr <- c(t(v1) %*% v2y, t(v1) %*% v2x) / (2 * (t(y) %*% v2x))  # center
    }

    return(list(ctr=ctr, rad=rad))
}

```

### Vertex that produces the circle with the maximum radius

Used later in `getMinCircle()`

```{r}
getMaxRad <- function(xy, S) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) >= 2, ncol(xy) == 2)
    stopifnot(is.numeric(S), length(S) >= 2, length(S) <= nrow(xy))

    n    <- length(S)                    # number of points
    Sidx <- seq(along=S)                 # index for points
    rads <- numeric(n)                   # radii for all circles
    post <- (Sidx %% n) + 1              # next point in S
    prev <- Sidx[order(post)]            # previous point in S
    for(i in Sidx) {
        pts     <- rbind(xy[S[prev[i]], ], xy[S[i], ], xy[S[post[i]], ])
        rads[i] <- getCircleFrom3(pts)$rad  # circle radius
    }

    return(which.max(rads))
}
```

### Check if the angle at B in triangle ABC exceeds 90 degrees

```{r}
isBiggerThan90 <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) == 3, ncol(xy) == 2)
    d   <- dist(xy)
    dAB <- d[1]
    dAC <- d[2]
    dBC <- d[3]
    return((dAB^2 + dBC^2 - dAC^2) < 0)
}
```

### Maximum pairwise distance between two 2D-points

```{r}
getMaxPairDist <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), ncol(xy) == 2, nrow(xy) >= 2)

    # 2D -> only convex hull is relevant
    H    <- chull(xy)      # convex hull indices (vertices ordered clockwise)
    pts  <- xy[H, ]        # points that make up the convex hull
    N    <- nrow(pts)                      # number of points on hull
    dMat <- dist(pts, method="euclidean")  # distance matrix
    idx  <- which.max(as.matrix(dMat))     # maximum distance
    i    <- (idx-1) %/% N+1                # column -> point 1
    j    <- (idx-1) %%  N+1                # row    -> point 2
    mPts <- H[c(i, j)]                     # rows with max distance
    dst  <- max(dMat)                      # max distance

    return(list(d=dst, idx=mPts))
}
```

### Minimal enclosing circle

```{r}
getMinCircle <- function(xy) {
    stopifnot(is.matrix(xy), is.numeric(xy), nrow(xy) >= 2, ncol(xy) == 2)

    H    <- chull(xy)      # convex hull indices (vertices ordered clockwise)
    hPts <- xy[H, ]        # points that make up the convex hull

    ## remove collinear points on convex hull, if any
    del <- idCollinear(hPts)
    if(length(del) >= 1L) {
        H    <- H[-del]
        hPts <- hPts[-del, ]

        if(length(H) < 2L) {
            stop("less than 2 points left after removing collinear points on convex hull")
        }
    }

    ## min circle may touch convex hull in only two points
    ## if so, it is centered between the hull points with max distance
    maxPD  <- getMaxPairDist(hPts)
    idx    <- maxPD$idx    # index of points with max distance
    rad    <- maxPD$d / 2  # half the distance -> radius
    rangeX <- c(hPts[idx[1], 1], hPts[idx[2], 1])
    rangeY <- c(hPts[idx[1], 2], hPts[idx[2], 2])
    ctr    <- c(rangeX[1] + 0.5*diff(rangeX), rangeY[1] + 0.5*diff(rangeY))

    ## check if circle centered between hPts[pt1Idx, ] and hPts[pt2Idx, ]
    ## contains all points (all distances <= rad)
    dst2ctr <- dist(rbind(ctr, hPts[-idx, ]))      # distances to center
    if(all(as.matrix(dst2ctr)[-1, 1] <= rad)) {    # if all <= rad, we're done
        tri <- rbind(hPts[idx, ], ctr)
        return(getCircleFrom3(tri))
    }

    ## min circle touches hull in three points - Skyum algorithm
    S <- H                               # copy of hull indices that will be changed
    while(length(S) >= 2) {
        n    <- length(S)                # number of remaining hull vertices
        Sidx <- seq(along=S)             # index for vertices
        post <- (Sidx %% n) + 1          # next vertex in S
        prev <- Sidx[order(post)]        # previous vertex in S
        mIdx <- getMaxRad(xy, S)         # idx for maximum radius

        ## triangle where mIdx is vertex B in ABC
        Smax <- rbind(xy[S[prev[mIdx]], ], xy[S[mIdx], ], xy[S[post[mIdx]], ])

        ## if there's only two hull vertices, we're done
        if(n <= 2) { break }

        ## check if angle(ABC) is > 90
        ## if so, eliminate B - if not, we're done
        if(isBiggerThan90(Smax)) { S <- S[-mIdx] } else { break }
    }

    return(getCircleFrom3(Smax))
}
```

### Draw the minimal enclosing circle

```{r rerDiagBounding04}
mc     <- getMinCircle(xy)
angles <- seq(0, 2*pi, length.out=200)
circ   <- cbind(mc$ctr[1] + mc$rad*cos(angles),
                mc$ctr[2] + mc$rad*sin(angles))

# determine axis limits so that the circle will be visible
xLims <- mc$ctr[1] + c(-mc$rad, mc$rad)
yLims <- mc$ctr[2] + c(-mc$rad, mc$rad)
plot(xy, xlab="x", ylab="y", xlim=xLims, ylim=yLims, asp=1, type="n")
lines(circ, col="blue", lwd=2)
points(xy, pch=16, cex=1.5)
```

Notes
-------------------------

Note that the presented implementation of the minimum bounding box and of the minimum enclosing circle were written ad-hoc "to get the job done" with a limited amount of mostly random point data. The implementations are not rigorously tested for numerical stability, and are not optimized. Advanced implementations (with generalizations to higher dimensions) may be found in solid computer graphics libraries.

  * [Computational Geometry Algorithms Library. 2021. CGAL Chapter 65: Bounding Volumes](https://doc.cgal.org/Manual/latest/doc_html/cgal_manual/Bounding_volumes/Chapter_main.html)
  * [Fischer, K.; Gaertner, B.; Kutz, M. 2003. Fast smallest-enclosing-ball computation in high dimensions. In: Proceedings of the 11th  European Symposium on Algorithms (ESA), 630-641](https://github.com/hbf/miniball)
  * [Gaertner, B. 2021. Miniball: Smallest Enclosing Balls of Points](https://people.inf.ethz.ch/gaertner/subdir/software/miniball.html)
  * Skyum, S. 1991. A simple algorithm for computing the smallest enclosing circle. Information Processing Letters 37(3), 121-125.
  * Toussaint, G. T. 1983. Solving geometric problems with the rotating calipers. In: Proceedings of the 1983 IEEE MELECON. Athens, Greece: IEEE Computer Society.
  * Welzl, E. 1991. Smallest enclosing disks (balls and ellipsoids). In: Maurer H. (eds), New Results and New Trends in Computer Science 555, 359-370. doi 10.1007/BFb0038202.
  * Todd MJ and Yildrim EA. On Khachiyan's Algorithm for the Computation of Minimum Volume Enclosing Ellipsoids. Discrete Applied Mathematics 2007: 155: 1731-1744. [PDF](https://people.orie.cornell.edu/miketodd/TYKhach.pdf)
