---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Graphics devices: Opening and saving diagrams"
categories: [Diagrams, BasicDiagrams]
rerCat: Diagrams
tags: [Diagrams]
---

Graphics devices: Opening and saving diagrams
=========================

Install required packages
-------------------------

[`Cairo`](https://cran.r-project.org/package=Cairo)

```{r}
wants <- c("Cairo")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Device margins
-------------------------

![Device margins](http://www.dwoll.de/rexrepos/content/assets/img/dev_margins.svg)

See also the [RStudio cheat sheet: How big is your graph?](https://www.rstudio.com/resources/cheatsheets/)


Opening and closing a device
-------------------------

```{r}
dev.new(); dev.new(); dev.new()
dev.list()
dev.cur()
dev.set(3)
dev.set(dev.next())
dev.off()
graphics.off()
```

Saving plots to a graphics file
-------------------------

### Regular R device

`pdf()`, `png()` and `jpeg()` are functions that open a graphics file of the corresponding type. See `?device` for more.

```{r eval=FALSE}
pdf("pdf_test.pdf", width=5, height=5)
plot(1:10, rnorm(10))
dev.off()
```

```{r eval=FALSE}
plot(1:10, rnorm(10))
dev.copy(jpeg, filename="copied.jpg", quality=90)
graphics.off()
```

### Device from package `Cairo`

`Cairo()` has a `dpi` option for choosing the graphics resolution in the output file. Option `type` chooses the type of the graphics file.

```{r eval=FALSE}
library(Cairo)
Cairo(width=5, height=5, units="in", file="Cairo_pdf.pdf", type="pdf",
      bg="white", canvas="white", dpi=120)
plot(1:10, rnorm(10))
dev.off()
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:Cairo))
```
