---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Frequency tables"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

Frequency tables
=========================

Install required packages
-------------------------

[`DescTools`](https://cran.r-project.org/package=DescTools), [`dplyr`](https://cran.r-project.org/package=dplyr)

```{r}
wants <- c("DescTools", "dplyr")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Category frequencies for one variable
-------------------------

### Absolute frequencies

```{r}
set.seed(123)
(myLetters <- sample(LETTERS[1:5], 12, replace=TRUE))
(tab <- table(myLetters))
names(tab)
tab["B"]
```

```{r rerFrequencies01}
barplot(tab, main="Counts")
```

### (Cumulative) relative frequencies

```{r}
(relFreq <- prop.table(tab))
```

```{r}
cumsum(relFreq)
```

### Counting non-existent categories

```{r}
letFac <- factor(myLetters, levels=c(LETTERS[1:5], "Q"))
letFac
table(letFac)
```

Counting runs
-------------------------

```{r}
(vec <- rep(rep(c("f", "m"), 3), c(1, 3, 2, 4, 1, 2)))
```

```{r}
(res <- rle(vec))
```

```{r}
length(res$lengths)
```

```{r}
inverse.rle(res)
```

Contingency tables for two or more variables
-------------------------

### Absolute frequencies using `table()`

```{r}
N    <- 10
(sex <- factor(sample(c("f", "m"), N, replace=TRUE)))
(work <- factor(sample(c("home", "office"), N, replace=TRUE)))
(cTab <- table(sex, work))
```

```{r}
summary(cTab)
```

```{r rerFrequencies02}
barplot(cTab, beside=TRUE, legend.text=rownames(cTab), ylab="absolute frequency")
```

### Using `xtabs()`

```{r}
counts   <- sample(0:5, N, replace=TRUE)
(persons <- data.frame(sex, work, counts))
```

```{r}
xtabs(~ sex + work, data=persons)
xtabs(counts ~ sex + work, data=persons)
```

### Marginal sums and means

```{r}
rowSums(cTab)
colMeans(cTab)
addmargins(cTab, c(1, 2), FUN=mean)
```

### Relative frequencies

```{r}
(relFreq <- prop.table(cTab))
```

### Conditional relative frequencies

```{r}
prop.table(cTab, margin=1)
```

```{r}
prop.table(cTab, margin=2)
```

### Flat contingency tables for more than two variables

```{r}
(group <- factor(sample(c("A", "B"), 10, replace=TRUE)))
ftable(work, sex, group, row.vars="work", col.vars=c("sex", "group"))
```

Recovering the original data from contingency tables
-------------------------

Individual-level data frame

```{r}
library(DescTools)
Untable(cTab)
```

Group-level data frame

```{r}
as.data.frame(cTab, stringsAsFactors=TRUE)
```

Percentile rank
-------------------------

```{r}
(vec <- round(rnorm(10), 2))
library(DescTools)
100*PercentRank(vec)
```

Using base R

```{r}
Fn <- ecdf(vec)
Fn(vec)
100 * Fn(0.1)
Fn(sort(vec))
knots(Fn)
```

```{r rerFrequencies03}
plot(Fn, main="cumulative frequencies")
```

Using package `dplyr`
-------------------------

Data set

```{r}
N      <- 12
sex    <- factor(sample(c("f", "m"), N, replace=TRUE), levels=c("f", "m"))
group  <- factor(sample(rep(c("CG", "WL", "T"), 4), N, replace=FALSE), levels=c("CG", "WL", "T"))
age    <- sample(18:35, N, replace=TRUE)
IQ     <- round(rnorm(N, mean=100, sd=15))
rating <- round(runif(N, min=0, max=6))
myDf1  <- data.frame(id=1:N, sex, group, age, IQ, rating)
```

### Absolute frequencies 

`count()` gives frequencies using new variable `n` (by default).

```{r}
library(dplyr)
myDf1 %>%
    count(sex, group, name="n")
```

By default, `count()` drops groups / combinations of groups with no entries. Use option `.drop=FALSE` to include entries with frequency 0.

```{r}
myDf1 %>%
    count(sex, group, .drop=FALSE)
```

Add corresponding count to existing data frame.

```{r}
myDf1 %>%
    add_count(sex, group)
```

### Relative frequencies

```{r}
myDf1 %>%
    count(sex, group) %>%
    mutate(freq_rel=n / sum(n))

myDf1 %>%
    add_count(sex, group) %>%
    mutate(freq_rel=n / n())
```

### Conditional relative frequencies

```{r}
myDf1 %>%
    count(sex, group, name="n_sex_group") %>%
    group_by(sex) %>%
    mutate(n_sex=sum(n_sex_group),
           freq_cond_rel=n_sex_group / n_sex) %>%
    ungroup()
```

```{r}
myDf1 %>%
    add_count(sex, name="n_sex") %>%
    add_count(sex, group, name="n_sex_group") %>%
    mutate(freq_cond_rel=n_sex_group / n_sex) %>%
    select(id, sex, group, n_sex, n_sex_group, freq_cond_rel)
```

### Percent rank

```{r}
myDf1 %>%
    mutate(rating_pr=100*percent_rank(rating))
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:DescTools))
try(detach(package:dplyr))
```
