---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Generate systematic and random data"
categories: [RBasics]
rerCat: R_Basics
tags: [RBasics]
---

Generate systematic and random data
=========================

Empty objects
-------------------------

```{r}
N <- 5
numeric(N)
matrix(numeric(N))
character(N)
vector(mode="list", length=N)
```

Systematic numerical sequences
-------------------------

### Ordered sequences

```{r}
20:26
26:20
-4:2
-(4:2)
seq(from=2, to=12, by=2)
seq(from=2, to=11, by=2)
seq(from=0, to=-1, length.out=5)
```

```{r}
age <- c(18, 20, 30, 24, 23, 21)
seq(along=age)
vec <- numeric(0)
length(vec)
1:length(vec)
seq(along=vec)
```

### Repeated sequences

```{r}
rep(1:3, times=5)
rep(c("A", "B", "C"), times=c(2, 3, 4))
rep(age, each=2)
```

Random data
-------------------------

Strictly, the data is pseudorandom. There are several options for the random number generator, see `RNGkind()`. Use `set.seed()` to set the state of the RNG. This allows to replicate the following sequence of numbers. Copy `.Random.seed` into your own object to save the current state of the RNG. Don't modify `.Random.seed`.

### Sample from an urn

```{r}
set.seed(123)
sample(1:6, size=20, replace=TRUE)
sample(c("rot", "gruen", "blau"), size=8, replace=TRUE)
x <- c(2, 4, 6, 8)
sample(x[(x %% 4) == 0])
sample(x[(x %% 8) == 0])
```

### Data from random variables with different distributions

```{r}
runif(5, min=1, max=6)
rbinom(20, size=5, prob=0.3)
rchisq(4, df=7)
rnorm(6, mean=100, sd=15)
rt(5, df=5, ncp=1)
rf(5, df1=2, df2=10)
```

See `?Distributions` for more distribution types. Even more information can be found in CRAN task view [Probability Distributions](https://CRAN.R-project.org/view=Distributions).
