---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "R user interfaces"
categories: [RBasics]
rerCat: R_Basics
tags: [GUI, "Command line"]
---

R user interfaces
=========================

R's default user interface
----------------

### R console

Command prompt

```
>
```

For interactive data analysis: type in commands, hit `Return` key and get text output as a result.

```{r}
1+7
```

Comments start with `#`, they are not executed as a command

```{r}
9 * 3     # this is a comment
```

 * Copy and paste to and from the clipboard as usual (Windows: `Ctrl+c` and `Ctrl+v`)
 * Interrupt R with `ESC` (on Windows) or `Ctrl+c` (Linux)
 * Quit with `q()`
 * Save a protocol of your commands and the output with `sink("fileName.txt", split=TRUE)`

### Non-interactive use

For batch mode:

`Rscript.exe input.r`

`Rterm.exe --no-restore --no-save < input.r > output.txt`

### Get and set options

```{r}
getOption("width")
```

Change option, save previous value, and restore previous value

```{r}
op <- options(width=70)
options(op)
```

Also see `help(Startup)` for files controlling the startup options.

Contributed user interfaces to R
----------------

Compared to the standard user interface that is already included with R, there are several better alternative options.

### For working with R commands

 - [RStudio](http://www.rstudio.org/) integrated development environment: Cross platform (Windows, MacOS, Linux), great support for the [workflow for these posts](<%= @config[:rer][:baseurl] %>/rerWorkflow.html), my preferred choice
 - [Eclipse](http://www.eclipse.org/eclipse) integrated development environment with [StatET](http://www.walware.de/goto/statet) plugin: Cross platform (Windows, MacOS, Linux), powerful, visual debugging support, somewhat complicated to set up ([installation instructions](http://www.catherinedalzell.ca/wp-content/uploads/2012/09/Installing-R-Sweave-and-Eclipse-on-Windows.pdf)), somewhat sluggish on older computers
 - [Architect](http://www.openanalytics.eu/architect) by OpenAnalytics: Combines Eclipse and StatET into a pre-configured bundle that removes the complicated Eclipse configuration
 - [TinnR](http://sourceforge.net/projects/tinn-r) text editor with good support for communicating with R: Windows only
 - [Emacs](http://www.gnu.org/software/emacs/) / [XEmacs](http://www.xemacs.org/) text editor with [Emacs Speaks Statistics](http://ess.r-project.org/) add-on: Cross platform (Windows, MacOS, Linux), very powerful, somewhat hard to learn

### Graphical front-ends for R functions

 - [Rcmdr](http://socserv.mcmaster.ca/jfox/Misc/Rcmdr/): R Commander - A Basic-Statistics GUI for R based on Java
 - [RKWard](http://rkward.sourceforge.net/) graphical user interface to R: Linux and limited Windows support
