---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Getting help and documentation"
categories: [RBasics]
rerCat: R_Basics
tags: [Help, Documentation]
---

Getting help and documentation
=========================

R's own help system
-------------------------

### Help system for R functions

```{r eval=FALSE}
help.start()
help(round)
?round
?"/"
# not shown (opens browser window)
```

```{r}
args(round)
example(round)
```

### Getting help without knowing the function name

```{r eval=FALSE}
help.search("mean")
# not shown (opens browser window)
```

```{r}
apropos("mean")
```

Online documentation
-------------------------

### Search, mailing lists and Q&A sites

 * [R mailing lists](https://www.r-project.org/mail.html): Especially R-help, also see the [posting guide](https://www.r-project.org/posting-guide.html)
 * [Stack Overflow: R](https://stackoverflow.com/tags/R) Q&A site - the technical side of using R
 * [CrossValidated: R](https://stats.stackexchange.com/tags/R) Q&A site - the statistical side of using R

### Introductory websites and texts

 * [Quick-R](https://www.statmethods.net/)
 * [Cookbook for R](http://www.cookbook-r.com/)
 * [UCLA data analysis examples](https://stats.idre.ucla.edu/other/dae/)

### Official documentation

 * [R FAQ](https://cran.at.r-project.org/doc/FAQ/R-FAQ.html) (frequently asked questions)
 * [R for Windows FAQ](https://cran.at.r-project.org/bin/windows/base/rw-FAQ.html)
 * [R Installation and Administration](https://cran.at.r-project.org/doc/manuals/R-admin.html)
 * [R Data Import/Export](https://cran.at.r-project.org/doc/manuals/R-data.html)

Books
-------------------------

### Introductory statistics

 * Dalgaard, P. (2008). Introductory Statistics with R (2nd ed.). London, UK: Springer. [URL](http://www.biostat.ku.dk/~pd/ISwR.html)
 * Maindonald, J. & Braun, W. J. (2010). Data Analysis and Graphics Using R: An Example-Based Approach (3rd ed.). Cambridge, UK: Cambridge University Press. [URL](http://maths.anu.edu.au/~johnm/r-book/daagur3.html)
 * Verzani. J. (2014). Using R for Introductory Statistics (2nd ed.). Boca Raton, FL: Chapman & Hall/CRC.
 * Wollschlaeger, D. (2020). Grundlagen der Datenanalyse mit R (5th ed.). Berlin: Springer. [URL](http://www.dwoll.de/r/gddmr.php)
 
### Specialized and advanced statistical topics

  * Regressions models
      * Fox J, Weisberg S. 2019. An R Companion to Applied Regression (3rd ed). Thousand Oaks, CA: Sage. [URL](https://socialsciences.mcmaster.ca/jfox/Books/Companion/)
      * Fox J. 2020. Regression diagnostics (2nd ed). [URL](https://socialsciences.mcmaster.ca/jfox/Books/RegressionDiagnostics/)
	  * Harrell Jr FE. 2015. Regression Modeling Strategies (2nd ed). New York: Springer. [URL](http://biostat.mc.vanderbilt.edu/wiki/Main/RmS)
  * Multivariate analysis
      * Zelterman D. 2015. Applied multivariate statistics with R. New York, NY: Springer.
  * Linear mixed models
      * Galecki AT, Burzykowski T. 2013. Linear Mixed-Effects Models Using R: A Step-by-Step Approach. New York, NY: Springer.
      * Pinheiro JC, Bates, DM. 2000. Mixed-Effects Models in S and S-PLUS. New York, NY: Springer.
      * West BT, Welch, KB, Galecki AT. 2014. Linear mixed models: A practical guide using statistical software (2n ed). Boca Raton, FL: Chapman & Hall/CRC. [URL](http://www-personal.umich.edu/~bwest/almmussp.html)
  * Resampling methods
      * Chihara L, Hesterberg T. 2018. Mathematical Statistics with Resampling and R (2nd ed). Hoboken, NJ: Wiley. [URL](https://sites.google.com/site/chiharahesterberg/)
  * Time series
      * Shumway RH, Stoffer DS. 2016. Time series analysis and its applications (4th ed). New York, NY: Springer. [URL](https://www.stat.pitt.edu/stoffer/tsa4/)
      * Shumway RH, Stoffer DS. 2019. Time Series: A Data Analysis Approach Using R. Boca Raton, FL: Chapman & Hall/CRC. [URL](https://www.stat.pitt.edu/stoffer/tsda/)
      * Hyndman RJ, Athanasopoulos G. 2019. Forecasting: Principles and practice (2nd ed). Melbourne, Australia: OTexts. [URL](https://otexts.com/fpp2/)
  * Bayes methods
      * Kruschke JK. 2015. Doing Bayesian data analysis: A tutorial with R, JAGS, and Stan (2nd ed). Amsterdam: Academic Press. [URL](http://doingbayesiandataanalysis.blogspot.com/)
      * McElreath R. 2020. Statistical rethinking: A Bayesian course with examples in R and Stan (2nd ed). Boca Raton, FL: Chapman & Hall/CRC. [URL](https://xcelab.net/rm/statistical-rethinking/)
  * Spatial statistics
      * Bivand RS, Pebesma E, Gómez-Rubio V. 2013. Applied spatial data analysis with R (2nd ed). New York, NY: Springer. [URL](https://asdar-book.org/)
  * R for data science and machine learning:
      * James G, Witten D, Hastie T, Tibshirani R. 2013. An introduction to statistical learning with applications in R. New York, NY: Springer. [URL](http://faculty.marshall.usc.edu/gareth-james/ISL/)
	  * Kuhn M, Johnson K. 2013. Applied predictive modeling. New York, NY: Springer. [URL](http://appliedpredictivemodeling.com/)
      * Wickham H, Grolemund G. 2017. R for data science. Sebastopol, CA: O’Reilly. [URL](https://r4ds.had.co.nz/)

### Diagrams

 * Murrell, P. (2011). R Graphics (2nd ed.). Boca Raton, FL: Chapman & Hall/CRC. [URL](http://www.stat.auckland.ac.nz/~paul/RG2e/)
 * Sarkar, D. (2008). Lattice: Multivariate Data Visualization with R. New York, NY: Springer. [URL](http://lmdvr.r-forge.r-project.org/)
 * Wickham, H. (2009). ggplot2: Elegant Graphics for Data Analysis. New York, NY: Springer. [URL](http://ggplot2.org/book/)

### Programming with R

  * Chambers JM. 2016. Extending R. Boca Raton, FL: Chapman & Hall/CRC.
  * Gillespie C, Lovelace R. 2017. Efficient R programming. Sebastopol, CA: O’Reilly. [URL](https://csgillespie.github.io/efficientR/)
  * Wickham H. 2019. Advanced R (2nd ed). Boca Raton, FL: Chapman & Hall/CRC. [URL](https://adv-r.hadley.nz/)
  * Wickham H. 2015. R packages. Sebastopol, CA: O’Reilly. [URL](https://r-pkgs.org/)
 
### Transition from other statistical software packages

 * Muenchen, R. A. (2011). R for SAS and SPSS Users (2nd ed.). New York, NY: Springer. [URL](http://r4stats.com/)
 * Muenchen, R. A. & Hilbe, J. M. (2010). R for Stata Users. New York, NY: Springer. [URL](http://r4stats.com/)
 * [David Hiebeler: Matlab vs. R Reference](http://www.math.umaine.edu/~hiebeler/comp/matlabR.html)

### Dynamic documents and reproducible research

  * Xie Y. 2015. Dynamic documents with R and knitr (2nd ed). Boca Raton, FL: Chapman & Hall/CRC.
  * Xie Y, Dervieux C, Riederer E. 2020. R Markdown Cookbook. Boca Raton, FL: Chapman & Hall  CRC. [URL](https://bookdown.org/yihui/rmarkdown-cookbook/)
  * Xie Y, Allaire JJ, Grolemund G. 2018. R markdown: The definitive guide. Boca Raton, FL: Chapman &  Hall/CRC. [URL](https://bookdown.org/yihui/rmarkdown/)
 * Gandrud, C. (2020). Reproducible research with R & RStudio (3rd ed). Boca Raton, FL: Chapman & Hall/CRC. [URL](http://christophergandrud.github.io/RepResR-RStudio/)
 * Stodden, V., Leisch, F. & Peng, R. D. (2014). Implementing Reproducible Research. Boca Raton, FL: Chapman & Hall/CRC.

### Shiny for interactive Web-Apps:
  * Sievert C. 2020. Interactive web-based data visualization with R, plotly, and shiny. Boca Raton, FL: Chapman & Hall/CRC. [URL](https://plotly-r.com/)
  * Wickham H. 2020. Mastering shiny. Sebastopol, CA: O’Reilly. [URL](https://mastering-shiny.org/)
