---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Inter-rater-reliability and Intra-class-correlation"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

Inter-rater-reliability and Intra-class-correlation
=========================

Install required packages
-------------------------

[`DescTools`](https://cran.r-project.org/package=DescTools)

```{r}
wants <- c("DescTools")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Inter-rater-reliability
-------------------------

### Absolute agreement between two or more raters

#### Two raters

```{r}
categ <- c("V", "N", "P")
lvls  <- factor(categ, levels=categ)
rtr1  <- rep(lvls, c(60, 30, 10))
rtr2  <- rep(rep(lvls, nlevels(lvls)), c(53,5,2, 11,14,5, 1,6,3))
cTab  <- table(rtr1, rtr2)
addmargins(cTab)
```

```{r}
library(DescTools)
Agree(cbind(rtr1, rtr2))
```

#### Three raters

```{r}
rtr3 <- rep(rep(lvls, nlevels(lvls)), c(48,8,3, 15,10,7, 3,4,2))
Agree(cbind(rtr1, rtr2, rtr3))
```

### Cohen's unweighted \(\kappa\) for two raters

```{r}
cTab <- xtabs(~ rtr1 + rtr2)
library(DescTools)
CohenKappa(cTab, conf.conf.level=0.95)
```

### Cohen's weighted \(\kappa\) for two raters

Ordered categories

```{r}
categ <- c("<10%", "11-20%", "21-30%", "31-40%", "41-50%", ">50%")
lvls  <- factor(categ, levels=categ)
tv1   <- rep(lvls, c(22, 21, 23, 16, 10, 8))
tv2   <- rep(rep(lvls, nlevels(lvls)), c(5,8,1,2,4,2, 3,5,3,5,5,0, 1,2,6,11,2,1,
                                         0,1,5,4,3,3, 0,0,1,2,5,2, 0,0,1, 2,1,4))
cTab  <- table(tv1, tv2)
addmargins(cTab)
```

```{r}
cTab <- xtabs(~ tv1 + tv2)
library(DescTools)
CohenKappa(cTab, weights="Equal-Spacing", conf.level=0.95)
```

### Fleiss' \(\kappa\) for two or more raters

```{r}
rtr1 <- letters[c(4,2,2,5,2, 1,3,1,1,5, 1,1,2,1,2, 3,1,1,2,1, 5,2,2,1,1, 2,1,2,1,5)]
rtr2 <- letters[c(4,2,3,5,2, 1,3,1,1,5, 4,2,2,4,2, 3,1,1,2,3, 5,4,2,1,4, 2,1,2,3,5)]
rtr3 <- letters[c(4,2,3,5,2, 3,3,3,4,5, 4,4,2,4,4, 3,1,1,4,3, 5,4,4,4,4, 2,1,4,3,5)]
rtr4 <- letters[c(4,5,3,5,4, 3,3,3,4,5, 4,4,3,4,4, 3,4,1,4,5, 5,4,5,4,4, 2,1,4,3,5)]
rtr5 <- letters[c(4,5,3,5,4, 3,5,3,4,5, 4,4,3,4,4, 3,5,1,4,5, 5,4,5,4,4, 2,5,4,3,5)]
rtr6 <- letters[c(4,5,5,5,4, 3,5,4,4,5, 4,4,3,4,5, 5,5,2,4,5, 5,4,5,4,5, 4,5,4,3,5)]
ratings <- cbind(rtr1, rtr2, rtr3, rtr4, rtr5, rtr6)
```

```{r}
library(DescTools)
KappaM(ratings, conf.level=0.95)
```

### Kendall's \(W\) for continuous ordinal ratings and two or more raters

```{r}
rtr1 <- c(1, 6, 3, 2, 5, 4)
rtr2 <- c(1, 5, 6, 2, 4, 3)
rtr3 <- c(2, 3, 6, 5, 4, 1)
ratings <- cbind(rtr1, rtr2, rtr3)
```

```{r}
library(DescTools)
KendallW(ratings, test=TRUE)
```

Intra-class-correlation for interval-scale ratings and two or more raters
-------------------------

```{r}
rtr1 <- c(9, 6, 8, 7, 10, 6)
rtr2 <- c(2, 1, 4, 1,  5, 2)
rtr3 <- c(5, 3, 6, 2,  6, 4)
rtr4 <- c(8, 2, 8, 6,  9, 7)
ratings <- cbind(rtr1, rtr2, rtr3, rtr4)
```

```{r}
library(DescTools)
ICC(ratings)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:DescTools))
```
