---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Linear algebra calculations"
categories: [Multivariate]
rerCat: Multivariate
tags: [Multivariate]
---

Linear algebra calculations
=========================

Install required packages
-------------------------

[`expm`](http://cran.r-project.org/package=expm), [`mvtnorm`](http://cran.r-project.org/package=mvtnorm), [`pracma`](http://cran.r-project.org/package=pracma)

```{r}
wants <- c("expm", "mvtnorm", "pracma")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Matrix algebra
-------------------------

### Transpose

```{r}
N  <- 4
Q  <- 2
(X <- matrix(c(20, 26, 10, 19, 29, 27, 20, 12), nrow=N, ncol=Q))
t(X)
```

### Extracting the diagnoal and creating a diagonal matrix

Extract the diagonal of a given matrix.

```{r}
diag(cov(X))
```

Create a diagonal matrix based on a vector.

```{r}
diag(1:3)
diag(2)
diag(5, nrow=1)
```

### Multiplication

```{r}
(Xc <- diag(N) - matrix(rep(1/N, N^2), nrow=N))
(Xdot <- Xc %*% X)
(SSP <- t(Xdot) %*% Xdot)
crossprod(Xdot)
```

```{r}
(1/(N-1)) * SSP
(S <- cov(X))
Ds <- diag(1/sqrt(diag(S)))
Ds %*% S %*% Ds
cov2cor(S)
```

```{r}
b <- 2
a <- c(-2, 1)
sweep(b*X, 2, a, "+")
colLens <- sqrt(colSums(X^2))
sweep(X, 2, colLens, "/")
X %*% diag(1/colLens)
```

### Power

```{r}
B <- cbind(c(1,1,1), c(0,2,0), c(0,0,2))
B %*% B %*% B
library(expm)
B %^% 3
```

### Cross product

```{r}
a <- c(1, 2, 3)
b <- c(4, 5, 6)
library(pracma)
cross(a, b)
```

Solving linear equations and calculating the inverse
-------------------------

### Inverse

```{r}
Y     <- matrix(c(1, 1, 1, -1), nrow=2)
(Yinv <- solve(Y))
Y %*% Yinv
```

### Moore-Penrose generalized inverse

```{r}
library(MASS)
gInv <- ginv(X)
zapsmall(gInv %*% X)
```

### Solving linear equations

```{r}
A  <- matrix(c(9, 1, -5, 0), nrow=2)
b  <- c(5, -3)
(x <- solve(A, b))
A %*% x
```

Norms and distances of matrices and vectors
-------------------------

### Norm

```{r}
a1 <- c(3, 4, 1, 8, 2)
sqrt(crossprod(a1))
sqrt(sum(a1^2))
```

```{r}
a2 <- c(6, 9, 10, 8, 7)
A  <- cbind(a1, a2)
sqrt(diag(crossprod(A)))
sqrt(colSums(A^2))
```

```{r}
norm(A, type="F")
sqrt(crossprod(c(A)))
```

### Distance

Length of difference vector

```{r}
set.seed(123)
B <- matrix(sample(-20:20, 12, replace=TRUE), ncol=3)
sqrt(crossprod(B[1, ] - B[2, ]))
```

```{r}
dist(B, diag=TRUE, upper=TRUE)
```

### Mahalanobis-transformation

```{r}
library(mvtnorm)
N     <- 100
mu    <- c(-3, 2, 4)
sigma <- matrix(c(4,2,-3, 2,16,-1, -3,-1,9), byrow=TRUE, ncol=3)
Y     <- round(rmvnorm(N, mean=mu, sigma=sigma))
```

```{r}
ctr   <- colMeans(Y)
S     <- cov(Y)
Seig  <- eigen(S)
sqrtD <- sqrt(Seig$values)
SsqrtInv <- Seig$vectors %*% diag(1/sqrtD) %*% t(Seig$vectors)

Xdot  <- sweep(Y, 2, ctr, "-")
Xmt   <- t(SsqrtInv %*% t(Xdot))
zapsmall(cov(Xmt))
colMeans(Xmt)
```

### Mahalanobis-distance

```{r}
ideal <- c(1, 2, 3)
y1    <- Y[1, ]
y2    <- Y[2, ]
mat   <- rbind(y1, y2)
```

```{r}
mahalanobis(mat, ideal, S)
```

```{r}
Sinv <- solve(S)
t(y1-ideal) %*% Sinv %*% (y1-ideal)
t(y2-ideal) %*% Sinv %*% (y2-ideal)
```

```{r}
mDist <- mahalanobis(Y, ideal, S)
min(mDist)
(idxMin <- which.min(mDist))
Y[idxMin, ]
```

```{r}
idealM <- t(SsqrtInv %*% (ideal - ctr))
crossprod(Xmt[1, ] - t(idealM))
crossprod(Xmt[2, ] - t(idealM))
```

Trace, determinant, rank, null space, condition index
-------------------------

### Trace

```{r}
(A <- matrix(c(9, 1, 1, 4), nrow=2))
sum(diag(A))
sum(diag(t(A) %*% A))
sum(diag(A %*% t(A)))
sum(A^2)
```

### Determinant

```{r}
det(A)
B <- matrix(c(-3, 4, -1, 7), nrow=2)
all.equal(det(A %*% B), det(A) * det(B))
det(diag(1:4))
Ainv <- solve(A)
all.equal(1/det(A), det(Ainv))
```

### Rank

```{r}
qrA <- qr(A)
qrA$rank
```

```{r}
(eigA <- eigen(A))
zapsmall(eigA$vectors %*% t(eigA$vectors))
sum(eigA$values)
prod(eigA$values)
```

### Null space (kernel)

```{r}
library(MASS)
Xnull <- Null(X)
t(X) %*% Xnull
```

### Condition index

We need to specify `base::norm()` here because after attaching package `expm` above, there is another function present with the same name.

```{r}
X <- matrix(c(20, 26, 10, 19, 29, 27, 20, 12, 17, 23, 27, 25), nrow=4)
kappa(X, exact=TRUE)
Xplus <- solve(t(X) %*% X) %*% t(X)
base::norm(X, type="2") * base::norm(Xplus, type="2")
```

```{r}
evX <- eigen(t(X) %*% X)$values
sqrt(max(evX) / min(evX[evX >= .Machine$double.eps]))
sqrt(evX / min(evX[evX >= .Machine$double.eps]))
```

Matrix decompositions
-------------------------

### Eigenvalues and eigenvectors

```{r}
X  <- matrix(c(20, 26, 10, 19, 29, 27, 20, 12, 17, 23, 27, 25), nrow=4)
(S <- cov(X))
eigS <- eigen(S)
G    <- eigS$vectors
D    <- diag(eigS$values)
G %*% D %*% t(G)
```

### Singular value decomposition

```{r}
svdX <- svd(X)
all.equal(X, svdX$u %*% diag(svdX$d) %*% t(svdX$v))
all.equal(sqrt(eigen(t(X) %*% X)$values), svdX$d)
```

### Cholesky decomposition

```{r}
R <- chol(S)
all.equal(S, t(R) %*% R)
```

### \(QR\)-decomposition

We need to specify `base::qr.Q()` here because after attaching package `expm` above, there is another function present with the same name.

```{r}
qrX <- qr(X)
Q   <- base::qr.Q(qrX)
R   <- base::qr.R(qrX)
all.equal(X, Q %*% R)
```

### Square-root

```{r}
library(expm)
sqrtm(S)
```

```{r}
sqrtD <- diag(sqrt(eigS$values))
(A <- G %*% sqrtD %*% t(G))
A %*% A
```

### \(X = N N^{t}\)

```{r}
N <- eigS$vectors %*% sqrt(diag(eigS$values))
N %*% t(N)
```

Orthogonal projections
-------------------------

### Direct implementation of \((X^{t} X)^{-1} X^{t}\)

```{r}
X    <- matrix(c(20, 26, 10, 19, 29, 27, 20, 12, 17, 23, 27, 25), nrow=4)
ones <- rep(1, nrow(X))
P1   <- ones %*% solve(t(ones) %*% ones) %*% t(ones)
P1x  <- P1 %*% X
head(P1x)
```

```{r}
a  <- ones / sqrt(crossprod(ones))
P2 <- a %*% t(a)
all.equal(P1, P2)
```

```{r}
IP1  <- diag(nrow(X)) - P1
IP1x <- IP1 %*% X
all.equal(IP1x, sweep(X, 2, colMeans(X), "-"))
```

```{r}
A   <- cbind(c(1, 0, 0), c(0, 1, 0))
P3  <- A %*% solve(t(A) %*% A) %*% t(A)
Px3 <- t(P3 %*% t(X))
Px3[1:3, ]
```

### Numerically stable implementation using the \(QR\)-decomposition

We need to specify `base::qr.Q()` here because after attaching package `expm` above, there is another function present with the same name.

```{r}
qrX   <- qr(X)
Q     <- base::qr.Q(qrX)
R     <- base::qr.R(qrX)
Xplus <- solve(t(X) %*% X) %*% t(X)
all.equal(Xplus, solve(R) %*% t(Q))
all.equal(X %*% Xplus, tcrossprod(Q))
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:MASS))
try(detach(package:expm))
try(detach(package:Matrix))
try(detach(package:pracma))
try(detach(package:mvtnorm))
```
