---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "The mean and other location measures"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

The mean and other location measures
=========================

TODO
-------------------------

 - link to npWilcoxon for `wilcox.test()`

Install required packages
-------------------------

[`DescTools`](https://cran.r-project.org/package=DescTools), [`modeest`](https://cran.r-project.org/package=modeest), [`robustbase`](https://cran.r-project.org/package=robustbase)

```{r}
wants <- c("DescTools", "modeest", "robustbase")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Mean, weighted mean, geometric mean, harmonic mean, and mode
-------------------------

### Mean

```{r}
age <- c(17, 30, 30, 25, 23, 21)
mean(age)
```

### Weighted mean

```{r}
weights <- c(0.6, 0.6, 0.3, 0.2, 0.4, 0.6)
weighted.mean(age, weights)
```

### Geometric mean

```{r}
library(DescTools)
Gmean(age)
```

### Harmonic mean

```{r}
library(DescTools)
Hmean(age)
```

### Mode

```{r}
vec <- c(11, 22, 22, 33, 33, 33, 33)
library(modeest)
mfv(vec)
mlv(vec, method="mfv")
```

Robust location measures
-------------------------

### Median

```{r}
median(age)
```

### Quantiles

Default: Quartiles

```{r}
(quant <- quantile(age))
quant[c("25%", "50%")]
```

Set probabilities with option `probs`

```{r}
quantile(age, probs=c(1/3, 2/3))
```

### Trimmed mean

```{r}
mean(age, trim=0.2)
```

### Winsorized mean

```{r}
library(DescTools)
(ageWins <- Winsorize(age, probs=c(0.2, 0.8)))
mean(ageWins)
```

### Huber-\(M\) estimator

```{r}
library(robustbase)
hM <- huberM(age)
hM$mu
```

### Hodges-Lehmann estimator (pseudo-median)

```{r}
library(DescTools)
HodgesLehmann(age, conf.level=0.95)
```

### Hodges-Lehmann estimator of difference between two location parameters

```{r}
N <- 8
X <- rnorm(N, 100, 15)
Y <- rnorm(N, 110, 15)
wilcox.test(X, Y, conf.int=TRUE)$estimate
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:modeest))
try(detach(package:DescTools))
try(detach(package:robustbase))
```
