---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Deal with missing data"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

Deal with missing data
=========================

Identify missing data with `is.na()`
-------------------------

### Find out which and how many observations have missing data

In R, missing values are coded as `NA` (not available)

```{r}
vec <- c(2, 5, 7)
vec[5]
```

```{r}
(vec1 <- c(10, 20, NA, 40, 50, NA))
length(vec1)
```

Identify missing values with `is.na()`

```{r}
is.na(vec1)
any(is.na(vec1))
which(is.na(vec1))
sum(is.na(vec1))
```

### Identify missing data in matrices

```{r}
vec2   <- c(NA, 7, 9, 10, 1, 8)
(matNA <- rbind(vec1, vec2))
is.na(matNA)
```

Behavior of `NA` in different situations
-------------------------

### Missing data in index vectors

```{r}
LETTERS[c(1, NA, 3)]
```

### Missing data in factors

```{r}
factor(LETTERS[c(1, NA, 3)])
factor(LETTERS[c(1, NA, 3)], exclude=NULL)
```

### Missing data in logical expressions

```{r}
NA & TRUE
TRUE | NA
```

```{r}
vecNA   <- c(-3, 2, 0, NA, -7, 5)
(logIdx <- vecNA > 0)
vecNA[logIdx]
vecNA[which(logIdx)]
```

Code missing values as `NA`
-------------------------

When data is entered in other applications (spreadsheets, SPSS, etc.), missing values are often coded as a reserved numeric value, e.g., 99 or 9999. These values need to be replaced with `NA`.

### In vectors

```{r}
vec <- c(30, 25, 23, 21, -999, 999)
is.na(vec) <- vec %in% c(-999, 999)
vec
```

### In matrices

```{r}
(mat <- matrix(c(30, 25, 23, 21, -999, 999), nrow=2, ncol=3))
is.na(mat) <- mat %in% c(-999, 999)
mat
```

Statistical analysis with missing data
-------------------------

### In vectors

```{r}
vecNA <- c(-3, 2, 0, NA, -7, 5)
mean(vecNA)
```

```{r}
goodIdx <- !is.na(vecNA)
mean(vecNA[goodIdx])
sd(na.omit(vecNA))
sum(vecNA, na.rm=TRUE)
```

### In matrices

```{r}
ageNA  <- c(18, NA, 27, 22)
DV1    <- c(NA, 1, 5, -3)
DV2    <- c(9, 4, 2, 7)
(matNA <- cbind(ageNA, DV1, DV2))
```

```{r}
apply(matNA, 1, FUN=mean)
apply(matNA, 1, FUN=mean, na.rm=TRUE)
```

### Casewise deletion of missing data

```{r}
(rowNAidx <- apply(is.na(matNA), 1, any))
matNA[!rowNAidx, ]
```

```{r}
na.omit(matNA)
colMeans(na.omit(matNA))
```

```{r}
cov(matNA, use="complete.obs")
all(cov(matNA, use="complete.obs") == cov(na.omit(matNA)))
```

Set casewise deletion as a permanent option for statistical functions (another choice is `"na.fail"`)

```{r eval=FALSE}
options(na.action="na.omit")
```

### Pairwise deletion of missing data

```{r}
rowMeans(matNA)
rowMeans(mat, na.rm=TRUE)
cov(matNA, use="pairwise.complete.obs")
```

Further resources
-------------------------

  * [CRAN Task View 'Missing Data'](https://CRAN.R-project.org/view=MissingData)

Useful packages
-------------------------

Multiple imputation is supported by functions in packages [`mice`](https://cran.r-project.org/package=mice) and [`Amelia`](https://cran.r-project.org/package=Amelia).
