---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Principal components analysis"
categories: [Multivariate]
rerCat: Multivariate
tags: [PCA]
---

Principal components analysis
=========================

Install required packages
-------------------------

[`mvtnorm`](https://cran.r-project.org/package=mvtnorm), [`robustbase`](https://cran.r-project.org/package=robustbase), [`pcaPP`](https://cran.r-project.org/package=pcaPP)

```{r}
wants <- c("mvtnorm", "robustbase", "pcaPP")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

PCA
-------------------------
    
### Using `prcomp()`

```{r}
set.seed(123)
library(mvtnorm)
Sigma <- matrix(c(4, 2, 2, 3), ncol=2)
mu    <- c(1, 2)
N     <- 50
X     <- rmvnorm(N, mean=mu, sigma=Sigma)
```

```{r}
(pca <- prcomp(X))
```

```{r}
summary(pca)
pca$sdev^2 / sum(diag(cov(X)))
```

```{r rerMultPCA01}
plot(pca)
```

For rotated principal components, see `principal()` from package [`psych`](https://cran.r-project.org/package=psych).

### Using `princomp()`

```{r}
(pcaPrin <- princomp(X))
(G <- pcaPrin$loadings)
```

### Find principal components

Principal component values for original data.

```{r}
pcVal <- predict(pca)
head(pcVal, n=5)
```

Principal component values for new data.

```{r}
Xnew <- matrix(1:4, ncol=2)
predict(pca, newdata=Xnew)
```

### Illustration

```{r}
B    <- G %*% diag(pca$sdev)
ctr  <- colMeans(X)
xMat <- rbind(ctr[1] - B[1, ], ctr[1])
yMat <- rbind(ctr[2] - B[2, ], ctr[2])
ab1  <- solve(cbind(1, xMat[ , 1]), yMat[ , 1])
ab2  <- solve(cbind(1, xMat[ , 2]), yMat[ , 2])
```

```{r rerMultPCA02}
plot(X, xlab="x", ylab="y", pch=20, asp=1,
     main="Data und principal components")
abline(coef=ab1, lwd=2, col="gray")
abline(coef=ab2, lwd=2, col="gray")
matlines(xMat, yMat, lty=1, lwd=6, col="blue")
points(ctr[1], ctr[2], pch=16, col="red", cex=3)
legend(x="topleft", legend=c("data", "PC axes", "SDs of PC", "centroid"),
       pch=c(20, NA, NA, 16), lty=c(NA, 1, 1, NA), lwd=c(NA, 2, 2, NA),
       col=c("black", "gray", "blue", "red"), bg="white")
```

### Dimension reduction using principal components

#### Full reproduction using all principal components

```{r}
Xdot <- scale(X, center=TRUE, scale=FALSE)
Y    <- Xdot %*% G
B    <- G %*% diag(pca$sdev)
H    <- scale(Y)
HB   <- H %*% t(B)

repr <- sweep(HB, 2, ctr, "+")
all.equal(X, repr)
sum((X-repr)^2)
```

#### Dimension reduction based on approximation by only the first principal component

```{r}
HB1   <- H[ , 1, drop=FALSE] %*% t(B[ , 1, drop=FALSE])
repr1 <- sweep(HB1, 2, ctr, "+")
sum((X-repr1)^2)
qr(scale(repr1, center=TRUE, scale=FALSE))$rank
```

```{r rerMultPCA03}
plot(X, xlab="x", ylab="y", pch=20, asp=1, main="Data und approximation")
abline(coef=ab1, lwd=2, col="gray")
abline(coef=ab2, lwd=2, col="gray")
segments(X[ , 1], X[ , 2], repr1[ , 1], repr1[ , 2])
points(repr1, pch=1, lwd=2, col="blue", cex=2)
points(ctr[1], ctr[2], pch=16, col="red", cex=3)
legend(x="topleft", legend=c("data", "PC axes", "centroid", "approximation"),
       pch=c(20, NA, 16, 1), lty=c(NA, 1, NA, NA), lwd=c(NA, 2, NA, 2),
       col=c("black", "gray", "red", "blue"), bg="white")
```

### Approximate the covariance matrix using principal components

```{r}
B %*% t(B)
cov(X)
B[ , 1, drop=FALSE] %*% t(B[ , 1, drop=FALSE])
```

Robust PCA
-------------------------

```{r}
library(robustbase)
princomp(X, cov=covMcd(X))
```

```{r}
library(pcaPP)
PCAproj(X, k=ncol(X), method="qn")
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:pcaPP))
try(detach(package:mvtnorm))
try(detach(package:robustbase))
```
