---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Exploratory factor analysis for ordinal categorical data"
categories: [Multivariate]
rerCat: Multivariate
tags: [FactorAnalysis]
---

Exploratory factor analysis for ordinal categorical data
=========================

Install required packages
-------------------------

[`GPArotation`](http://cran.r-project.org/package=GPArotation), [`mvtnorm`](http://cran.r-project.org/package=mvtnorm), [`polycor`](http://cran.r-project.org/package=polycor), [`psych`](http://cran.r-project.org/package=psych)

```{r}
wants <- c("GPArotation", "mvtnorm", "polycor", "psych")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Factor analysis
-------------------------

### Simulate categorical data based on continuous variables

First, let's simulate 200 observations from 6 variables, coming from 2 orthogonal factors. I'll take a couple of intermediate steps and start with multivariate normal continuous data that I later dichotomize. That way, we can compare Pearson correlations with polychoric correlations, and compare factor loadings from continuous data with that from dichotomous data and the true loadings.

```{r}
set.seed(123)
N <- 200                       # number of observations
P <- 6                         # number of variables
Q <- 2                         # number of factors

# true P x Q loading matrix -> variable-factor correlations
Lambda <- matrix(c(0.7,-0.4, 0.8,0, -0.2,0.9, -0.3,0.4, 0.3,0.7, -0.8,0.1),
                 nrow=P, ncol=Q, byrow=TRUE)
```

Now simulate the actual data from the model \(x = \Lambda f + e\), with \(x\) being the observed variable values of a person, \(\Lambda\) the true loadings matrix, \(f\) the latent factor score, and \(e\) iid, mean 0, normal errors.

```{r}
# factor scores (uncorrelated factors)
library(mvtnorm)               # for rmvnorm()
FF <- rmvnorm(N, mean=c(5, 15), sigma=diag(Q))

# matrix with iid, mean 0, normal errors
E   <- rmvnorm(N, rep(0, P), diag(P))
X   <- FF %*% t(Lambda) + E    # matrix with variable values
dfX <- data.frame(X)           # data also as a data frame
```

Now let's categorize the data. We'll keep the data in two formats: as a data frame with ordered factors, and as a numeric matrix. `hetcor()` from package `polycor` gives us the polychoric correlation matrix we'll later use for the FA.

```{r}
# categorize variables into a list of ordered factors
lOrd <- lapply(dfX, function(x) {
               cut(x, breaks=quantile(x), include.lowest=TRUE,
                   ordered=TRUE, labels=LETTERS[1:4]) })
dfOrd  <- data.frame(lOrd)     # combine list into a data frame
ordNum <- data.matrix(dfOrd)   # categorized data as a numeric matrix
```

### Factor analysis for ordered categorical data

Use the polychoric correlation matrix to do a regular FA.

```{r}
library(polycor)               # for hetcor()
pc <- hetcor(dfOrd, ML=TRUE)   # polychoric corr matrix
```

```{r}
library(psych)
faPC <- fa(r=pc$correlations, nfactors=2, n.obs=N, rotate="varimax")
faPC$loadings
```

It is possible to skip the step of calculating the polychoric correlation matrix, and directly use `fa.poly()` from package `psych`, which does the same thing in the end. This function accepts the raw dichotomous data as a numeric matrix.

```{r results='hide'}
# polychoric FA
faPCdirect <- fa.poly(ordNum, nfactors=2, rotate="varimax")
```

```{r}
faPCdirect$fa$loadings         # loadings are the same as above ...
```

Factor scores
-------------------------

For factor scores, look at package `ltm` which has a `factor.scores()` function specifically for polytomous outcome data. An example is provided on [this page](http://eur.academia.edu/DimitrisRizopoulos/Teaching) -> "Factor Scores - Ability Estimates".

Visualize loadings
-------------------------

You can visualize the loadings from the factor analysis using `factor.plot()` and `fa.diagram()`, both from package `psych`. For some reason, `factor.plot()` accepts only the `$fa` component of the result from `fa.poly()`, not the full object.

```{r rerMultFApoly01}
factor.plot(faPCdirect$fa, cut=0.5)
fa.diagram(faPCdirect)
```

Determine number of factors
-------------------------

Parallel analysis and a "very simple structure" analysis provide help in selecting the number of factors. Again, package `psych` has the required functions. `vss()` takes the polychoric correlation matrix as an argument.

```{r rerMultFApoly02, results='hide'}
fap <- fa.parallel.poly(ordNum)   # parallel analysis for dichotomous data
```

```{r rerMultFApoly03}
fap
vss(pc$correlations, n.obs=N, rotate="varimax")  # very simple structure
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:GPArotation))
try(detach(package:psych))
try(detach(package:polycor))
try(detach(package:sfsmisc))
try(detach(package:mvtnorm))
```

(based on an answer I wrote on CrossValidated)
