---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Exploratory factor analysis"
categories: [Multivariate]
rerCat: Multivariate
tags: [FactorAnalysis]
---

Exploratory factor analysis
=========================

Install required packages
-------------------------

[`GPArotation`](https://cran.r-project.org/package=GPArotation), [`mvtnorm`](https://cran.r-project.org/package=mvtnorm), [`psych`](https://cran.r-project.org/package=psych)

```{r}
wants <- c("GPArotation", "mvtnorm", "psych")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Factor analysis
-------------------------

### Simulate data

True matrix of loadings

```{r}
N <- 200
P <- 6
Q <- 2
(Lambda <- matrix(c(0.7,-0.4, 0.8,0, -0.2,0.9, -0.3,0.4, 0.3,0.7, -0.8,0.1),
                  nrow=P, ncol=Q, byrow=TRUE))
```

Non correlated factors

```{r}
set.seed(123)
library(mvtnorm)
Kf <- diag(Q)
mu <- c(5, 15)
FF <- rmvnorm(N, mean=mu,        sigma=Kf)
E  <- rmvnorm(N, mean=rep(0, P), sigma=diag(P))
X  <- FF %*% t(Lambda) + E
```

### Using `factanal()`

```{r}
(fa <- factanal(X, factors=2, scores="regression"))
```

### Using `fa()` from package `psych` with rotation

Rotation uses package `GPArotation`

```{r}
library(psych)
corMat <- cor(X)
(faPC  <- fa(r=corMat, nfactors=2, n.obs=N, rotate="varimax"))
```

Factor scores
-------------------------

```{r}
bartlett <- fa$scores
head(bartlett)
```

```{r}
anderson <- factor.scores(x=X, f=faPC, method="Anderson")
head(anderson$scores)
```

Visualize loadings
-------------------------

```{r rerMultFA01}
factor.plot(faPC, cut=0.5)
fa.diagram(faPC)
```

Determine number of factors
-------------------------

Parallel analysis and a "very simple structure" analysis provide help in selecting the number of factors. Again, package `psych` has the required functions. `vss()` takes the polychoric correlation matrix as an argument.

```{r rerMultFA02}
fa.parallel(X)                     # parallel analysis
vss(X, n.obs=N, rotate="varimax")  # very simple structure
```

Useful packages
-------------------------

For confirmatory factor analysis (CFA), see packages [`sem`](https://cran.r-project.org/package=sem), [`OpenMx`](https://openmx.ssri.psu.edu/), and [`lavaan`](https://cran.r-project.org/package=lavaan) which implement structural equation models. More packages can be found in CRAN task view [Psychometric Models](https://CRAN.R-project.org/view=Psychometrics).

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:psych))
try(detach(package:mvtnorm))
```
