---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Assess univariate and multivariate normality"
categories: [Univariate]
rerCat: Univariate
tags: [ANOVA, tTests]
---

Assess univariate and multivariate normality
=========================

Install required packages
-------------------------

[`energy`](http://cran.r-project.org/package=energy), [`ICS`](http://cran.r-project.org/package=ICS), [`mvtnorm`](http://cran.r-project.org/package=mvtnorm)

```{r}
wants <- c("energy", "ICS", "mvtnorm")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Univariate normality
-------------------------

### QQ-plot

```{r rerNormality01}
set.seed(123)
DV <- rnorm(20, mean=1.5, sd=3)
qqnorm(DV, pch=20, cex=2)
qqline(DV, col="gray60", lwd=2)
```

### Shapiro-Wilk-test

Composite null hypothesis: any normal distribution

```{r}
shapiro.test(DV)
```

### Anderson-Darling-test

Composite null hypothesis: any normal distribution

```{r}
library(DescTools)
AndersonDarlingTest(DV)
```

### Cramer-von-Mises-test

Composite null hypothesis: any normal distribution

```{r}
library(DescTools)
CramerVonMisesTest(DV)
```

### Shapiro-Francia-test

Composite null hypothesis: any normal distribution

```{r}
library(DescTools)
ShapiroFranciaTest(DV)
```

### Jarque-Bera-test

Composite null hypothesis: any normal distribution

```{r}
library(DescTools)
JarqueBeraTest(DV)
```

### Kolmogorov-Smirnov-test

Exact null hypothesis: fully specified normal distribution

```{r}
ks.test(DV, "pnorm", mean=1, sd=2, alternative="two.sided")
```

### Lilliefors-test

Composite null hypothesis: any normal distribution

```{r}
library(DescTools)
LillieTest(DV)
```

### Pearson \(\chi^{2}\)-test

Tests weaker null hypothesis (any distribution with the same probabilities for the given class intervals).

Wrong: `pearson.test()` does not use grouped ML-estimate or maximum \(\chi^{2}\)-estimate

```{r}
library(DescTools)
PearsonTest(DV, n.classes=6, adjust=TRUE)
```

Multivariate normality
-------------------------

### Energy-test

```{r}
mu    <- c(2, 4, 5)
Sigma <- matrix(c(4,2,-3, 2,16,-1, -3,-1,9), byrow=TRUE, ncol=3)
library(mvtnorm)
X <- rmvnorm(100, mu, Sigma)
```

```{r}
library(energy)                    # for mvnorm.etest()
mvnorm.etest(X)
```

### Kurtosis- and skew-test

#### Kurtosis-test

```{r}
library(ICS)
mvnorm.kur.test(X)
```

#### Skew-test
```{r}
library(ICS)
X <- rmvnorm(100, c(2, 4, 5))
mvnorm.skew.test(X)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:DescTools))
try(detach(package:energy))
try(detach(package:ICS))
try(detach(package:mvtnorm))
try(detach(package:survey))
try(detach(package:CompQuadForm))
```
