---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Nonparametric location tests for one and two samples"
categories: [Nonparametric, ClassicalNonparametric]
rerCat: Nonparametric
tags: [Nonparametric, ClassicalNonparametric]
---

Nonparametric location tests for one and two samples
=========================

Install required packages
-------------------------

[`coin`](https://cran.r-project.org/package=coin), [`DescTools`](https://cran.r-project.org/package=DescTools)

```{r}
wants <- c("coin", "DescTools")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

One-sample
-------------------------

### Sign-test

Two-sided test

```{r}
set.seed(123)
medH0 <- 30
DV    <- sample(0:100, 20, replace=TRUE)

library(DescTools)
SignTest(DV, mu=medH0)
```

### Wilcoxon signed rank test

```{r}
IQ    <- c(99, 131, 118, 112, 128, 136, 120, 107, 134, 122)
medH0 <- 110
```

```{r}
wilcox.test(IQ, alternative="greater", mu=medH0, conf.int=TRUE)
```

Two independent samples
-------------------------

### Sign-test

```{r}
Nj  <- c(20, 30)
DVa <- rnorm(Nj[1], mean= 95, sd=15)
DVb <- rnorm(Nj[2], mean=100, sd=15)
wIndDf <- data.frame(DV=c(DVa, DVb),
                     IV=factor(rep(1:2, Nj), labels=LETTERS[1:2]))
```

Looks at the number of cases in each group which are below or above the median of the combined data.

```{r}
library(coin)
median_test(DV ~ IV, distribution="exact", data=wIndDf)
```

### Wilcoxon rank-sum test (\(=\) Mann-Whitney \(U\)-test)

```{r}
wilcox.test(DV ~ IV, alternative="less", conf.int=TRUE, data=wIndDf)
```

```{r}
library(coin)
wilcox_test(DV ~ IV, alternative="less", conf.int=TRUE,
            distribution="exact", data=wIndDf)
```

### van der Waerden normal scores test

```{r}
library(coin)
normal_test(DV ~ IV, distribution=approximate(nresample=9999),
            data=wIndDf)
```

### Fisher-Pitman permutation test with untransformed response values

```{r}
library(coin)
oneway_test(DV ~ IV, distribution=approximate(nresample=9999),
            data=wIndDf)
```

Two dependent samples
-------------------------

### Sign-test

```{r}
N      <- 20
DVpre  <- rnorm(N, mean= 95, sd=15)
DVpost <- rnorm(N, mean=100, sd=15)
wDepDf <- data.frame(id=factor(rep(1:N, times=2)),
                     DV=c(DVpre, DVpost),
                     IV=factor(rep(0:1, each=N), labels=c("pre", "post")))
```

Two-sided test

```{r}
medH0  <- 0
DVdiff <- aggregate(DV ~ id, FUN=diff, data=wDepDf)

library(DescTools)
SignTest(DVdiff$DV, mu=medH0)
```

### Wilcoxon signed rank test

```{r}
library(coin)
wilcoxsign_test(DV ~ IV | id, alternative="less",
                distribution="exact", data=wDepDf)
```

### van der Waerden normal scores test

```{r}
library(coin)
normal_test(DV ~ IV | id, alternative="less",
            distribution=approximate(nresample=9999), data=wDepDf)
```

### Fisher-Pitman permutation test with untransformed response values

```{r}
library(coin)
oneway_test(DV ~ IV | id, distribution=approximate(nresample=9999),
            data=wDepDf)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:DescTools))
try(detach(package:coin))
try(detach(package:survival))
```
