---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Integration and Differentiation"
categories: [Numerical]
rerCat: Numerical
tags: [MDS]
---

Integration and Differentiation
=========================

Install required packages
-------------------------

[`numDeriv`](https://cran.r-project.org/package=numDeriv)

```{r}
wants <- c("numDeriv")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Integration
-------------------------

### Example

Half the normal density is to the left of the mean

```{r}
integrate(dnorm, -Inf, 1, mean=1, sd=2)$value
pnorm(1, mean=1, sd=2)
```

### Re-implement normal cumulative distribution function

Case study for implementing a cumulative distribution function which may not be available in closed form whereas the corresponding density function is.

Naive implementation of normal density - numerically unstable, only works within narrow parameter range.

```{r}
dGauss <- function(x, mu=0, sigma=1) {
    (1/(sigma*sqrt(2*pi))) * exp(-0.5 * ((x-mu)/sigma)^2)
}
```

Implement normal cumulative distribution function by integrating density function.

```{r}
pGauss <- function(x, mu=0, sigma=1) {
    integrate(dGauss, lower=-Inf, upper=x, mu=mu, sigma=sigma)$value
}
```

Check custom CDF against R implementation

```{r}
pGauss(1, mu=0, sigma=1)
pnorm(1, mean=0, sd=1)
```

Differentiation
-------------------------

### Examples

Density function = derivative of cumulative distribution function. Check with normal distribution.

```{r}
library(numDeriv)
x <- seq(-2, 2, length.out=5)
grad(pnorm, x, mean=0, sd=1)
dnorm(x, mean=0, sd=1)
```

Derivative of exponential function is exponential function.

```{r}
library(numDeriv)
exp(x)
grad(exp, x)
```

### Delta method for standard error of function of parameter estimates

Maximum likelihood parameter estimate in Poisson regression

```{r}
N  <- 100
X  <- rnorm(N, 0, 2)
mu <- exp(1 + 0.5*X)
Y  <- rpois(N, mu)
glmFit <- glm(Y ~ X, family=poisson(link="log"))
(bML   <- cbind(coef(glmFit)))
```

Function of the first two parameter estimates -> Evaluate function at ML estimate.

```{r}
f <- function(b) { b[1] / b[2] }
f(bML)
```

Delta method -> approximate standard error of function of parameter estimates -> approximate Wald confidence interval

```{r}
library(numDeriv)
(bGrad <- grad(f, bML))
(SEdelta <- sqrt(t(bGrad) %*% vcov(glmFit) %*% bGrad))
(CIdelta <- c(lo=f(bML) - c(qnorm(1-(0.05/2))*SEdelta),
              up=f(bML) + c(qnorm(1-(0.05/2))*SEdelta)))
```

### Standard errors from Hessian matrix

Poisson negative log-likelihood

```{r}
nllPois <- function(b, X, Y) {
    mu <- exp(b[1] + b[2]*X)
    -sum(Y*log(mu) - mu)
}
```

Inverse of Hessian at ML estimate = variance-covariance matrix of parameter estimates.

```{r}
h <- hessian(nllPois, x=bML, X=X, Y=Y)
solve(h)
```

Compare to `vcov()`

```{r}
vcov(glmFit)
```

Further resources
-------------------------

  * [CRAN Task View 'Numerical Mathematics'](https://CRAN.R-project.org/view=NumericalMathematics)

Useful packages
-------------------------

  * [`statmod`](https://cran.r-project.org/package=statmod) - `gauss.quad()` for Gaussian quadrature
  * [`cubature`](https://cran.r-project.org/package=cubature)
  * [`polyCub`](https://cran.r-project.org/package=polyCub)

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:numDeriv))
```
