---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Multinomial regression"
categories: [Univariate, Regression]
rerCat: Univariate
tags: [Regression, GLM]
---

Multinomial regression
=========================

TODO
-------------------------

 - link to regressionLogistic, regressionOrdinal

Install required packages
-------------------------

[`mlogit`](http://cran.r-project.org/package=mlogit), [`nnet`](http://cran.r-project.org/package=nnet), [`VGAM`](http://cran.r-project.org/package=VGAM)

```{r}
wants <- c("mlogit", "nnet", "VGAM")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Multinomial regression
----------------------

### Simulate data
    
```{r}
set.seed(123)
N      <- 100
X1     <- rnorm(N, 175, 7)
X2     <- rnorm(N,  30, 8)
Ycont  <- 0.5*X1 - 0.3*X2 + 10 + rnorm(N, 0, 6)
Ycateg <- cut(Ycont, breaks=quantile(Ycont), include.lowest=TRUE,
              labels=c("--", "-", "+", "++"), ordered=FALSE)
dfMN   <- data.frame(X1, X2, Ycateg)
```

### Using `vglm()` from package `VGAM`

Estimator based on likelihood-inference

```{r}
library(VGAM)
vglmFitMN <- vglm(Ycateg ~ X1 + X2, family=multinomial(refLevel=1), data=dfMN)
```

Odds ratios

```{r}
exp(VGAM::coef(vglmFitMN))
```

### Using `multinom()` from package `nnet`

Estimator based on neural networks -> slightly different results than `vglm()`, `mlogit()`

```{r results='hide'}
library(nnet)
(mnFit <- multinom(Ycateg ~ X1 + X2, data=dfMN))
# not shown
```

### Using `mlogit()` from package `mlogit`

Uses person-choice (long) format, so data frame has to be reshaped with `mlogit.data()`

```{r results='hide'}
library(mlogit)
dfMNL <- mlogit.data(dfMN, choice="Ycateg", shape="wide", varying=NULL)
(mlogitFit <- mlogit(Ycateg ~ 0 | X1 + X2, reflevel="--", data=dfMNL))
# not shown
```

Predicted category membership
-------------------------

### Predicted category probabilities

```{r}
PhatCateg <- VGAM::predict(vglmFitMN, type="response")
head(PhatCateg)
```

```{r results='hide'}
predict(mnFit, type="probs")
fitted(mlogitFit, outcome=FALSE)
# not shown
```

### Predicted categories

```{r}
categHat <- levels(dfMN$Ycateg)[max.col(PhatCateg)]
head(categHat)
```

```{r results='hide'}
predCls <- predict(mnFit, type="class")
head(predCls)
# not shown
```

Assess model fit
-------------------------

### Classification table

```{r}
facHat <- factor(categHat, levels=levels(dfMN$Ycateg))
cTab   <- xtabs(~ Ycateg + facHat, data=dfMN)
addmargins(cTab)
```

Correct classification rate

```{r}
(CCR <- sum(diag(cTab)) / sum(cTab))
```

### Deviance, log-likelihood and AIC

```{r}
VGAM::deviance(vglmFitMN)
VGAM::logLik(vglmFitMN)
VGAM::AIC(vglmFitMN)
```

### McFadden, Cox & Snell and Nagelkerke pseudo \(R^{2}\)

Log-likelihoods for full model and 0-model without predictors X1, X2

```{r}
vglm0 <- vglm(Ycateg ~ 1, family=multinomial(refLevel=1), data=dfMN)
LLf   <- VGAM::logLik(vglmFitMN)
LL0   <- VGAM::logLik(vglm0)
```

McFadden pseudo-\(R^2\)

```{r}
as.vector(1 - (LLf / LL0))
```

Cox & Snell

```{r}
as.vector(1 - exp((2/N) * (LL0 - LLf)))
```

Nagelkerke

```{r}
as.vector((1 - exp((2/N) * (LL0 - LLf))) / (1 - exp(LL0)^(2/N)))
```

Apply regression model to new data
-------------------------

### Simulate new data

`predict.mlogit()` requires a new data frame in long format. Therefore also add new (irrelevant) categorical responses to enable reshaping the data frame with `mlogit.data()`.

```{r}
Nnew  <- 3
dfNew <- data.frame(X1=rnorm(Nnew, 175, 7),
                    X2=rnorm(Nnew,  30, 8),
                    Ycateg=factor(sample(c("--", "-", "+", "++"), Nnew, TRUE),
                                  levels=c("--", "-", "+", "++")))
```

### Predicted class probabilities

```{r}
VGAM::predict(vglmFitMN, dfNew, type="response")
```

```{r results='hide'}
predict(mnFit, dfNew, type="probs")

dfNewL <- mlogit.data(dfNew, choice="Ycateg", shape="wide", varying=NULL)
predict(mlogitFit, dfNewL)
# not shown
```

Coefficient tests and overall model test
-------------------------

### Individual coefficient tests

Estimated standard deviations, z-values, and p-values for parameters based on normality on assumption that z-values are asymptotically \(N(0, 1)\) distributed.

```{r}
sumMN   <- VGAM::summary(vglmFitMN)
(coefMN <- VGAM::coef(sumMN))
```

Approximative Wald-based confidence intervals

```{r}
zCrit   <- qnorm(c(0.05/2, 1 - 0.05/2))
(ciCoef <- t(apply(coefMN, 1, function(x) x["Estimate"] - zCrit*x["Std. Error"] )))
```

Tests for other models.

```{r results='hide'}
summary(mnFit)
summary(mlogitFit)
# not shown
```

### Model comparisons - likelihood-ratio tests

Likelihood-ratio-test for predictor `X2`

We need to specify `VGAM::lrtest()` here because after attaching package `mlogit` above, there is another function present with the same name.

```{r}
vglmFitR <- vglm(Ycateg ~ X1, family=multinomial(refLevel=1), data=dfMN)
VGAM::lrtest(vglmFitMN, vglmFitR)
```

Likelihood-ratio-test for the full model against the 0-model without predictors (just intercept)

```{r}
VGAM::lrtest(vglmFitMN, vglm0)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:mlogit))
try(detach(package:Formula))
try(detach(package:maxLik))
try(detach(package:miscTools))
try(detach(package:nnet))
try(detach(package:VGAM))
try(detach(package:splines))
try(detach(package:stats4))
```
