---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Ordinal regression"
categories: [Univariate, Regression]
rerCat: Univariate
tags: [Regression, GLM]
---

Ordinal regression
=========================

```{r echo=FALSE}
library(knitr)
opts_knit$set(self.contained=FALSE)
opts_chunk$set(tidy=FALSE, message=FALSE, warning=FALSE, comment=NA)
```

TODO
-------------------------

 - link to regressionLogistic, regressionMultinom

Install required packages
-------------------------

[`MASS`](https://cran.r-project.org/package=MASS), [`ordinal`](https://cran.r-project.org/package=ordinal), [`rms`](https://cran.r-project.org/package=rms), [`VGAM`](https://cran.r-project.org/package=VGAM), [`DescTools`](https://cran.r-project.org/package=DescTools)

```{r}
wants <- c("MASS", "ordinal", "rms", "VGAM", "DescTools")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Ordinal regression (proportional odds model)
-------------------------
    
### Simulate data

Dependent variable \(Y_{\text{ord}}\) with \(k=4\) groups, \(p=2\) predictor variables

```{r}
set.seed(123)
N     <- 100
X1    <- rnorm(N, 175, 7)
X2    <- rnorm(N,  30, 8)
Ycont <- 0.5*X1 - 0.3*X2 + 10 + rnorm(N, 0, 6)
Yord  <- cut(Ycont, breaks=quantile(Ycont), include.lowest=TRUE,
             labels=c("--", "-", "+", "++"), ordered=TRUE)
dfOrd <- data.frame(X1, X2, Yord)
```

### Using `vglm()` from package `VGAM`

Model using cumulative logits: \(\text{logit}(p(Y \geq g)) = \ln \frac{P(Y \geq g)}{1 - P(Y \geq g)} = \beta_{0_{g}} + \beta_{1} X_{1} + \dots + \beta_{p} X_{p} \quad(g = 2, \ldots, k)\)

```{r}
library(VGAM)
(vglmFit <- vglm(Yord ~ X1 + X2, family=propodds, data=dfOrd, model=TRUE))
```

Equivalent:

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=cumulative(parallel=TRUE, reverse=TRUE), data=dfOrd)
# not shown
```

Adjacent category logits \(\ln \frac{P(Y=g)}{P(Y=g-1)}\) with proportional odds assumption

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=acat(parallel=TRUE), data=dfOrd)
# not shown
```

Continuation ratio logits \(\ln \frac{P(Y=g)}{P(Y < g)}\) with proportional odds assumption (discrete version of Cox proportional hazards model for survival data)

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=sratio(parallel=TRUE), data=dfOrd)
# not shown
```

### Using `orm()` from package `rms`

Model \(\text{logit}(p(Y \geq g)) = \beta_{0_{g}} + \beta_{1} X_{1} + \dots + \beta_{p} X_{p} \quad(g = 2, \ldots, k)\)

```{r}
library(rms)
(ormFit <- orm(Yord ~ X1 + X2, data=dfOrd))
```

### Using `polr()` from package `MASS`

Model \(\text{logit}(p(Y \leq g)) = \beta_{0_{g}} - (\beta_{1} X_{1} + \dots + \beta_{p} X_{p}) \quad(g = 1, \ldots, k-1)\)

```{r results='hide'}
library(MASS)
(polrFit <- polr(Yord ~ X1 + X2, method="logistic", data=dfOrd))
# not shown
```

Profile likelihood based confidence intervals (need to use `MASS:::confint.polr()` instead of `confint()` since other packages are loaded, and method is masked).

```{r}
exp(MASS:::confint.polr(polrFit))
```

### Using `clm()` from package `ordinal`

Model \(\text{logit}(p(Y \leq g)) = \beta_{0_{g}} - (\beta_{1} X_{1} + \dots + \beta_{p} X_{p}) \quad(g = 1, \ldots, k-1)\)

```{r results='hide'}
library(ordinal)
(clmFit <- clm(Yord ~ X1 + X2, link="logit", data=dfOrd))
# not shown
```

Predicted category membership
-------------------------

### Predicted category probabilities

```{r}
PhatCateg <- VGAM::predict(vglmFit, type="response")
head(PhatCateg)
```

```{r results='hide'}
predict(ormFit, type="fitted.ind")
predict(clmFit, subset(dfOrd, select=c("X1", "X2"), type="prob"))$fit
predict(polrFit, type="probs")
# not shown
```

### Predicted categories

```{r}
categHat <- levels(dfOrd$Yord)[max.col(PhatCateg)]
head(categHat)
```

```{r results='hide'}
predict(clmFit, type="class")
predict(polrFit, type="class")
# not shown
```

Apply regression model to new data
-------------------------

### Simulate new data

```{r}
Nnew  <- 3
dfNew <- data.frame(X1=rnorm(Nnew, 175, 7),
                    X2=rnorm(Nnew,  30, 8))
```

### Predicted class probabilities

```{r}
VGAM::predict(vglmFit, dfNew, type="response")
```

```{r results='hide'}
predict(ormFit,  dfNew, type="fitted.ind")
predict(polrFit, dfNew, type="probs")
predict(clmFit,  subset(dfNew, select=c("X1", "X2"), type="prob"))$fit
# not shown
```

Assess model fit
-------------------------

### Classification table

```{r}
facHat <- factor(categHat, levels=levels(dfOrd$Yord))
cTab   <- xtabs(~ Yord + facHat, data=dfOrd)
addmargins(cTab)
```

Correct classification rate

```{r}
(CCR <- sum(diag(cTab)) / sum(cTab))
```

### Deviance, log-likelihood and AIC

```{r}
VGAM::deviance(vglmFit)
VGAM::logLik(vglmFit)
VGAM::AIC(vglmFit)
```

### McFadden, Cox & Snell and Nagelkerke pseudo \(R^{2}\)

```{r}
library(DescTools)
PseudoR2(vglmFit, which=c("McFadden", "CoxSnell", "Nagelkerke"))
```

Coefficient tests and overall model test
-------------------------

### Individual coefficient tests

Estimated standard deviations, z-values and p-values for parameters based on assumption that z-values are asymptotically \(N(0, 1)\) distributed.

```{r}
sumOrd   <- summary(vglmFit)
(coefOrd <- coef(sumOrd))
```

Profile likelihood confidence intervals

```{r}
VGAM::confint(vglmFit, method="profile")
```

Tests for other models.

```{r eval=FALSE}
summary(polrFit)
summary(clmFit)
# not shown
```

### Model comparisons - likelihood-ratio tests

Likelihood-ratio-test for predictor `X2`

```{r}
vglmR <- vglm(Yord ~ X1, family=propodds, data=dfOrd)
anova(vglmFit, vglmR, type="I")
```

Likelihood-ratio-test for the full model against the 0-model without predictors (just intercept)

```{r}
vglm0 <- vglm(Yord ~ 1, family=propodds, data=dfOrd)
anova(vglmFit, vglm0, type="I")
```

Test assumption of proportional odds (=parallel logits)
-------------------------

### Using `vglm()` from package `VGAM`

```{r}
vglmP  <- vglm(Yord ~ X1 + X2, family=cumulative(parallel=TRUE,  reverse=TRUE),
              data=dfOrd)

# vglmNP <- vglm(Yord ~ X1 + X2, family=cumulative(parallel=FALSE, reverse=TRUE),
#                data=dfOrd)
# VGAM::lrtest(vglmP, vglmNP)
```

### Using `clm()` from package `ordinal`

```{r}
clmP  <- clm(Yord ~ X1 + X2, link="logit", data=dfOrd)

## model with non-proportional odds for X2:
clmNP <- clm(Yord ~ X1, nominal=~X2, data=dfOrd)
anova(clmP, clmNP)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:VGAM))
try(detach(package:splines))
try(detach(package:stats4))
try(detach(package:rms))
try(detach(package:Hmisc))
try(detach(package:lattice))
try(detach(package:survival))
try(detach(package:Formula))
try(detach(package:ggplot2))
try(detach(package:MASS))
try(detach(package:ordinal))
try(detach(package:DescTools))
```
