---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Ordinal regression"
categories: [Univariate, Regression]
rerCat: Univariate
tags: [Regression, GLM]
---

Ordinal regression
=========================

```{r echo=FALSE}
library(knitr)
opts_knit$set(self.contained=FALSE)
opts_chunk$set(tidy=FALSE, message=FALSE, warning=FALSE, comment=NA)
# render_jekyll()
```

TODO
-------------------------

 - link to regressionLogistic, regressionMultinom

Install required packages
-------------------------

[`MASS`](http://cran.r-project.org/package=MASS), [`ordinal`](http://cran.r-project.org/package=ordinal), [`rms`](http://cran.r-project.org/package=rms), [`VGAM`](http://cran.r-project.org/package=VGAM)

```{r}
wants <- c("MASS", "ordinal", "rms", "VGAM")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Ordinal regression (proportional odds model)
-------------------------
    
### Simulate data

Dependent variable \(Y_{\text{ord}}\) with \(k=4\) groups, \(p=2\) predictor variables

```{r regressionOrdinalDf, cache=TRUE}
set.seed(123)
N     <- 100
X1    <- rnorm(N, 175, 7)
X2    <- rnorm(N,  30, 8)
Ycont <- 0.5*X1 - 0.3*X2 + 10 + rnorm(N, 0, 6)
Yord  <- cut(Ycont, breaks=quantile(Ycont), include.lowest=TRUE,
             labels=c("--", "-", "+", "++"), ordered=TRUE)
dfOrd <- data.frame(X1, X2, Yord)
```

### Using `vglm()` from package `VGAM`

Model using cumulative logits: \(\text{logit}(p(Y \geq g)) = \ln \frac{P(Y \geq g)}{1 - P(Y \geq g)} = \beta_{0_{g}} + \beta_{1} X_{1} + \dots + \beta_{p} X_{p} \quad(g = 2, \ldots, k)\)

```{r}
library(VGAM)
(vglmFit <- vglm(Yord ~ X1 + X2, family=propodds, data=dfOrd))
```

Equivalent:

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=cumulative(parallel=TRUE, reverse=TRUE), data=dfOrd)
# not shown
```

Adjacent category logits \(\ln \frac{P(Y=g)}{P(Y=g-1)}\) with proportional odds assumption

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=acat(parallel=TRUE), data=dfOrd)
# not shown
```

Continuation ratio logits \(\ln \frac{P(Y=g)}{P(Y < g)}\) with proportional odds assumption (discrete version of Cox proportional hazards model for survival data)

```{r results='hide'}
vglm(Yord ~ X1 + X2, family=sratio(parallel=TRUE), data=dfOrd)
# not shown
```

### Using `lrm()` from package `rms`

Model \(\text{logit}(p(Y \geq g)) = \beta_{0_{g}} + \beta_{1} X_{1} + \dots + \beta_{p} X_{p} \quad(g = 2, \ldots, k)\)

```{r}
library(rms)
(lrmFit <- lrm(Yord ~ X1 + X2, data=dfOrd))
```

### Using `polr()` from package `MASS`

Model \(\text{logit}(p(Y \leq g)) = \beta_{0_{g}} - (\beta_{1} X_{1} + \dots + \beta_{p} X_{p}) \quad(g = 1, \ldots, k-1)\)

```{r regressionOrdinalPolrFit, results='hide', cache=TRUE}
library(MASS)
(polrFit <- polr(Yord ~ X1 + X2, method="logistic", data=dfOrd))
# not shown
```

Profile likelihood based confidence intervals

```{r dependson=c("regressionOrdinalDf", "regressionOrdinalPolrFit")}
exp(confint(polrFit))
```

### Using `clm()` from package `ordinal`

Model \(\text{logit}(p(Y \leq g)) = \beta_{0_{g}} - (\beta_{1} X_{1} + \dots + \beta_{p} X_{p}) \quad(g = 1, \ldots, k-1)\)

```{r results='hide'}
library(ordinal)
(clmFit <- clm(Yord ~ X1 + X2, link="logit", data=dfOrd))
# not shown
```

Predicted category membership
-------------------------

### Predicted category probabilities

```{r}
PhatCateg <- predict(vglmFit, type="response")
head(PhatCateg)
```

```{r results='hide'}
predict(lrmFit, type="fitted.ind")
predict(clmFit, subset(dfOrd, select=c("X1", "X2"), type="prob"))$fit
predict(polrFit, type="probs")
# not shown
```

### Predicted categories

```{r}
categHat <- levels(dfOrd$Yord)[max.col(PhatCateg)]
head(categHat)
```

```{r results='hide'}
predict(clmFit, type="class")
predict(polrFit, type="class")
# not shown
```

Apply regression model to new data
-------------------------

### Simulate new data

```{r}
Nnew  <- 3
dfNew <- data.frame(X1=rnorm(Nnew, 175, 7),
                    X2=rnorm(Nnew,  30, 8))
```

### Predicted class probabilities

```{r}
predict(vglmFit, dfNew, type="response")
```

```{r results='hide'}
predict(lrmFit,  dfNew, type="fitted.ind")
predict(polrFit, dfNew, type="probs")
predict(clmFit,  subset(dfNew, select=c("X1", "X2"), type="prob"))$fit
# not shown
```

Assess model fit
-------------------------

### Classification table

```{r}
facHat <- factor(categHat, levels=levels(dfOrd$Yord))
cTab   <- table(dfOrd$Yord, facHat, dnn=c("Yord", "facHat"))
addmargins(cTab)
```

Correct classification rate

```{r}
(CCR <- sum(diag(cTab)) / sum(cTab))
```

### Deviance, log-likelihood and AIC

```{r}
deviance(vglmFit)
logLik(vglmFit)
AIC(vglmFit)
```

### McFadden, Cox & Snell and Nagelkerke pseudo \(R^{2}\)

Log-likelihoods for full model and 0-model without predictors X1, X2

```{r}
vglm0 <- vglm(Yord ~ 1, family=propodds, data=dfOrd)
LLf   <- logLik(vglmFit)
LL0   <- logLik(vglm0)
```

McFadden pseudo-\(R^2\)

```{r}
as.vector(1 - (LLf / LL0))
```

Cox & Snell

```{r}
as.vector(1 - exp((2/N) * (LL0 - LLf)))
```

Nagelkerke

```{r}
as.vector((1 - exp((2/N) * (LL0 - LLf))) / (1 - exp(LL0)^(2/N)))
```

Coefficient tests and overall model test
-------------------------

### Individual coefficient tests

Estimated standard deviations and z-values for parameters

```{r}
sumOrd   <- summary(vglmFit)
(coefOrd <- coef(sumOrd))
```

Approximative Wald-based confidence intervals

```{r}
zCrit   <- qnorm(c(0.05/2, 1 - 0.05/2))
(ciCoef <- t(apply(coefOrd, 1, function(x) x["Estimate"] - zCrit*x["Std. Error"] )))
```

p-values for two-sided paramter tests based on assumption that z-values are asymptotically \(N(0, 1)\) distributed

```{r}
2*(1 - pnorm(abs(coefOrd[ , "z value"])))
```

```{r results='hide', dependson=c("regressionOrdinalDf", "regressionOrdinalPolrFit")}
summary(polrFit)
summary(clmFit)
# not shown
```

### Model comparisons - likelihood-ratio tests

Likelihood-ratio-test for predictor `X2`

We need to specify `VGAM::lrtest()` here because after attaching package `mlogit` above, there is another function present with the same name.

```{r}
vglmR <- vglm(Yord ~ X1, family=propodds, data=dfOrd)
VGAM::lrtest(vglmFit, vglmR)
```

Likelihood-ratio-test for the full model against the 0-model without predictors (just intercept)

```{r}
VGAM::lrtest(vglmFit, vglm0)
```

Test assumption of proportional odds (=parallel logits)

```{r}
vglmP <- vglm(Yord ~ X1 + X2, family=cumulative(parallel=TRUE,  reverse=TRUE),
              data=dfOrd)

vglmNP <- vglm(Yord ~ X1 + X2, family=cumulative(parallel=FALSE, reverse=TRUE),
                data=dfOrd)
VGAM::lrtest(vglmP, vglmNP)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:ordinal))
try(detach(package:ucminf))
try(detach(package:Matrix))
try(detach(package:lattice))
try(detach(package:MASS))
try(detach(package:rms))
try(detach(package:Hmisc))
try(detach(package:survival))
try(detach(package:VGAM))
try(detach(package:splines))
try(detach(package:stats4))
```
