---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Poisson-regression"
categories: [Univariate, Regression]
rerCat: Univariate
tags: [Regression, GLM]
---

Poisson-regression
=========================

```{r echo=FALSE}
library(knitr)
opts_knit$set(self.contained=FALSE)
opts_chunk$set(tidy=FALSE, message=FALSE, warning=FALSE, comment=NA)
# render_jekyll()
```

Install required packages
-------------------------

[`lmtest`](http://cran.r-project.org/package=lmtest), [`MASS`](http://cran.r-project.org/package=MASS), [`mvtnorm`](http://cran.r-project.org/package=mvtnorm), [`pscl`](http://cran.r-project.org/package=pscl), [`sandwich`](http://cran.r-project.org/package=sandwich), [`VGAM`](http://cran.r-project.org/package=VGAM)

```{r}
wants <- c("lmtest", "MASS", "mvtnorm", "pscl", "sandwich", "VGAM")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Poisson regression
-------------------------
    
### Simulate data
    
```{r}
library(mvtnorm)
set.seed(123)
N     <- 200
sigma <- matrix(c(4,2,-3, 2,16,-1, -3,-1,8), byrow=TRUE, ncol=3)
mu    <- c(-3, 2, 4)
XY    <- rmvnorm(N, mean=mu, sigma=sigma)
Y     <- round(XY[ , 3] - 1.5)
Y[Y < 0] <- 0
dfCount <- data.frame(X1=XY[ , 1], X2=XY[ , 2], Y)
```

### Using `glm()`

```{r}
glmFitP <- glm(Y ~ X1 + X2, family=poisson(link="log"), data=dfCount)
summary(glmFitP)
```

Change factors for rate parameter \(\lambda\)

```{r}
exp(coef(glmFitP))
```

Profile likelihood based confidence intervals for change factors

```{r}
exp(confint(glmFitP))
```

### Using `vglm()` from package `VGAM`

```{r results='hide'}
library(VGAM)
summary(vglmFit <- vglm(Y ~ X1 + X2, family=poissonff, data=dfCount))
# not shown
```

### Analyse event rates

`offset` is the exposure \(\ln(t)\)

```{r}
Nt   <- 100
Ti   <- sample(20:40, Nt, replace=TRUE)
Xt   <- rnorm(Nt, 100, 15)
Yt   <- rbinom(Nt, size=Ti, prob=0.5)
glm(Yt ~ Xt, family=poisson(link="log"), offset=log(Ti))
```

Overdispersion
-------------------------

### Adjusted Poisson-regression

Same parameter estimates as in Poisson model, but different standard errors, hence different p-values

```{r}
glmFitQP <- glm(Y ~ X1 + X2, family=quasipoisson(link="log"), data=dfCount)
summary(glmFitQP)
```

Using `vglm()` from package `VGAM`

```{r results='hide'}
library(VGAM)
vglm(Y ~ X1 + X2, family=quasipoissonff, data=dfCount)
# not shown
```

### Heteroscedasticity consistent standard errors

Same parameter estimates as in Poisson model, but different standard errors, hence different p-values

```{r}
library(sandwich)
hcSE <- vcovHC(glmFitP, type="HC0")

library(lmtest)
coeftest(glmFitP, vcov=hcSE)
```

### Negative binomial regression

Using `glm.nb()` from package `MASS`

```{r}
library(MASS)
glmFitNB <- glm.nb(Y ~ X1 + X2, data=dfCount)
summary(glmFitNB)
```

Using `vglm()` from package `VGAM`

```{r results='hide'}
library(VGAM)
vglm(Y ~ X1 + X2, family=negbinomial, data=dfCount)
# not shown
```

### Test the negative binomial model against the Poisson model

```{r}
library(pscl)
odTest(glmFitNB)
```

Zero-inflated Regression models
-------------------------

### Zero-inflated Poisson regression

```{r}
library(pscl)
ziFitP <- zeroinfl(Y ~ X1 + X2 | 1, dist="poisson", data=dfCount)
summary(ziFitP)
```

Using `vglm()` from package `VGAM`

```{r results='hide'}
library(VGAM)
vglm(Y ~ X1 + X2, family=zipoissonff, data=dfCount)
# not shown
```

Vuong-Test using `vuong()` from package `pscl`: Poisson model against zero-inflated Poisson model

```{r}
library(pscl)
vuong(ziFitP, glmFitP)
```

### Zero-inflated negative binomial regression

```{r}
ziFitNB <- zeroinfl(Y ~ X1 + X2 | 1, dist="negbin", data=dfCount)
summary(ziFitNB)
```

Using `vglm()` from package `VGAM`

```{r results='hide'}
library(VGAM)
vglm(Y ~ X1 + X2, family=zinegbinomial, data=dfCount)
# not shown
```

Vuong-Test using `vuong()` from package `pscl`: negative binomial model against zero-inflated negative binomial model

```{r}
library(pscl)
vuong(ziFitNB, glmFitNB)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:VGAM))
try(detach(package:sandwich))
try(detach(package:lmtest))
try(detach(package:zoo))
try(detach(package:pscl))
try(detach(package:mvtnorm))
try(detach(package:lattice))
try(detach(package:splines))
try(detach(package:stats4))
try(detach(package:MASS))
```
