---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Binary logistic regression"
categories: [Univariate, Regression]
rerCat: Univariate
tags: [Regression, GLM]
---

Binary logistic regression
=========================

TODO
-------------------------

 - link to associationOrder for `pROC`, regressionOrdinal, regressionMultinom, regressionDiag for outliers, collinearity, crossvalidation

Install required packages
-------------------------

[`rms`](https://cran.r-project.org/package=rms), [`DescTools`](https://cran.r-project.org/package=DescTools)

```{r}
wants <- c("rms", "DescTools")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Descriptive model fit
-------------------------

### Simulate data
    
```{r}
set.seed(123)
SSRIpre  <- c(18, 16, 16, 15, 14, 20, 14, 21, 25, 11)
SSRIpost <- c(12,  0, 10,  9,  0, 11,  2,  4, 15, 10)
PlacPre  <- c(18, 16, 15, 14, 20, 25, 11, 25, 11, 22)
PlacPost <- c(11,  4, 19, 15,  3, 14, 10, 16, 10, 20)
WLpre    <- c(15, 19, 10, 29, 24, 15,  9, 18, 22, 13)
WLpost   <- c(17, 25, 10, 22, 23, 10,  2, 10, 14,  7)

P        <- 3
Nj       <- rep(length(SSRIpre), times=P)
IV       <- factor(rep(1:P, Nj), labels=c("SSRI", "Placebo", "WL"))
DVpre    <- c(SSRIpre,  PlacPre,  WLpre)
DVpost   <- c(SSRIpost, PlacPost, WLpost)
postFac  <- cut(DVpost, breaks=c(-Inf, median(DVpost), Inf),
                        labels=c("lo", "hi"))
dfAncova <- data.frame(IV, DVpre, DVpost, postFac)
```

```{r rerRegressionLogistic01}
cdplot(postFac ~ DVpre, data=dfAncova, subset=IV == "SSRI",
       main="Estimated categ probs SSRI")
cdplot(postFac ~ DVpre, data=dfAncova, subset=IV == "Placebo",
       main="Estimated categ probs placebo")
cdplot(postFac ~ DVpre, data=dfAncova, subset=IV == "WL",
       main="Estimated categ probs WL")
```

### Fit the model

```{r}
(glmFit <- glm(postFac ~ DVpre + IV, family=binomial(link="logit"), data=dfAncova))
```

Odds ratios

```{r}
exp(coef(glmFit))
```

Profile likelihood based confidence intervals for odds ratios

```{r}
exp(confint(glmFit))
```

### Fit the model based on a matrix of counts

```{r}
N      <- 100
x1     <- rnorm(N, 100, 15)
x2     <- rnorm(N, 10, 3)
total  <- sample(40:60, N, replace=TRUE)
hits   <- rbinom(N, total, prob=0.4)
hitMat <- cbind(hits, total-hits)
glm(hitMat ~ x1 + x2, family=binomial(link="logit"))
```

### Fit the model based on relative frequencies

```{r}
relHits <- hits/total
glm(relHits ~ x1 + x2, weights=total, family=binomial(link="logit"))
```

### Fitted logits and probabilities

```{r rerRegressionLogistic02}
logitHat <- predict(glmFit, type="link")
plot(logitHat, pch=16, col=c("red", "blue")[unclass(dfAncova$postFac)])
abline(h=0)
```

```{r}
Phat <- fitted(glmFit)
Phat <- predict(glmFit, type="response")
head(Phat)
mean(Phat)
prop.table(xtabs(~ postFac, data=dfAncova))
```

Assess model fit
-------------------------

### Classification table

```{r}
thresh <- 0.5
facHat <- cut(Phat, breaks=c(-Inf, thresh, Inf), labels=c("lo", "hi"))
cTab   <- xtabs(~ postFac + facHat, data=dfAncova)
addmargins(cTab)
```

Correct classification rate

```{r}
(CCR <- sum(diag(cTab)) / sum(cTab))
```

### log-Likelihood, AUC, Somers' \(D_{xy}\), Nagelkerke's pseudo \(R^{2}\)

Deviance, log-likelihood and AIC

```{r}
deviance(glmFit)
logLik(glmFit)
AIC(glmFit)
```

Nagelkerke's pseudo-\(R^{2}\) (R2), area under the ROC-Kurve (C), Somers' \(D_{xy}\) (Dxy), Goodman & Kruskal's \(\gamma\) (Gamma), Kendall's \(\tau\) (Tau-a)

```{r}
library(rms)
lrm(postFac ~ DVpre + IV, data=dfAncova)
```

For plotting the ROC-curve, see `pROC` in associationOrder

### McFadden, Cox & Snell and Nagelkerke pseudo \(R^{2}\)

```{r}
library(DescTools)
PseudoR2(glmFit, which=c("McFadden", "Nagelkerke", "CoxSnell"))
```

### Crossvalidation

`cv.glm()` function from package `boot`, see crossvalidation

### Apply model to new data

```{r}
Nnew  <- 3
dfNew <- data.frame(DVpre=rnorm(Nnew, 20, sd=7),
                    IV=factor(rep("SSRI", Nnew), levels=levels(dfAncova$IV)))
predict(glmFit, newdata=dfNew, type="response")
```

Coefficient tests and overall model test
-------------------------

### Individual coefficient tests

Wald-tests for parameters

```{r}
summary(glmFit)
```

Or see `lrm()` above

### Model comparisons - likelihood-ratio tests

```{r}
glm0 <- glm(postFac ~ 1, family=binomial(link="logit"), data=dfAncova)
anova(glm0, glmFit, test="Chisq")
```

```{r}
drop1(glmFit, test="Chi")
```

Or see `lrm()` above

### Model comparisons for testing IV

```{r}
glmPre <- update(glmFit, . ~ . - IV) # no IV factor
anova(glmPre, glmFit, test="Chisq")
```

### Model comparisons for testing DVpre

```{r}
anova(glm0, glmPre, test="Chisq")
```

Further resources
-------------------------

For penalized logistic regression, see packages [`logistf`](https://cran.r-project.org/package=logistf) (using Firth's penalized likelihood) and [`glmnet`](https://cran.r-project.org/package=glmnet). An example using `glmnet` for linear regression is in regressionRobPen.

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:rms))
try(detach(package:Hmisc))
try(detach(package:lattice))
try(detach(package:survival))
try(detach(package:Formula))
try(detach(package:ggplot2))
try(detach(package:DescTools))
```
