---
layout: post
title: "Survival analysis: Kaplan-Meier"
categories: [Univariate, Survival]
rerCat: Univariate
tags: [Survival, KaplanMeier]
---

Survival analysis: Kaplan-Meier
=========================

TODO
-------------------------

 - link to survivalCoxPH, survivalParametric

Install required packages
-------------------------

[`survival`](http://cran.r-project.org/package=survival)

```{r}
wants <- c("survival")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Simulated right-censored event times with Weibull distribution
-------------------------

Simulated survival time \(T\) influenced by time independent covariates \(X_{j}\) with effect parameters \(\beta_{j}\) under assumption of proportional hazards, stratified by sex.

\(T = (-\ln(U) \, b \, e^{-\bf{X} \bf{\beta}})^{\frac{1}{a}}\), where \(U \sim \mathcal{U}(0, 1)\), \(a\) is the Weibull shape parameter and \(b\) is the Weibull scale parameter.

```{r}
set.seed(1.234)
N      <- 180                  # number of observations
P      <- 3                    # number of groups
sex    <- factor(sample(c("f", "m"), N, replace=TRUE))  # stratification factor
X      <- rnorm(N, 0, 1)       # continuous covariate
IV     <- factor(rep(LETTERS[1:P], each=N/P))  # factor covariate
IVeff  <- c(0, -1, 1.5)        # effects of factor levels (1 -> reference level)
Xbeta  <- 0.7*X + IVeff[unclass(IV)] + rnorm(N, 0, 2)
weibA  <- 1.5                  # Weibull shape parameter
weibB  <- 100                  # Weibull scale parameter
U      <- runif(N, 0, 1)       # uniformly distributed RV
eventT <- ceiling((-log(U)*weibB*exp(-Xbeta))^(1/weibA))   # simulated event time

# censoring due to study end after 120 days
obsLen <- 120                  # length of observation time
censT  <- rep(obsLen, N)       # censoring time = end of study
obsT   <- pmin(eventT, censT)  # observed censored event times
status <- eventT <= censT      # has event occured?
dfSurv <- data.frame(obsT, status, sex, X, IV)          # data frame
```

Plot simulated data

```{r rerSurvivalKM01}
plot(ecdf(eventT), xlim=c(0, 200), main="Cumulative survival distribution",
     xlab="t", ylab="F(t)", cex.lab=1.4)
abline(v=obsLen, col="blue", lwd=2)
text(obsLen-5, 0.2, adj=1, labels="end of study", cex=1.4)
```

Kaplan-Meier-Analysis
-------------------------

### Estimate survival-function

Survival function \(S(t) = 1-F(t)\), hazard function

$$
\begin{equation*}
\lambda(t) = \lim_{\Delta_{t} \to 0^{+}} \frac{P(t \leq T < t + \Delta_{t} \, | \, T \geq t)}{\Delta_{t}} = \frac{P(t \leq T < t + \Delta_{t}) / \Delta_{t}}{P(T > t)} = \frac{f(t)}{S(t)}, \quad t \geq 0
\end{equation*}
$$

```{r}
library(survival)                # for Surv(), survfit()
## global estimate
KM0 <- survfit(Surv(obsT, status) ~ 1,  type="kaplan-meier", conf.type="log", data=dfSurv)

## separate estimate for all strata
(KM <- survfit(Surv(obsT, status) ~ IV, type="kaplan-meier", conf.type="log", data=dfSurv))
```

Arbitrary quantiles for estimated survival function.

```{r}
## survival 2.37-2 has a bug in quantile(), so this currently doesn't work
# quantile(KM0, probs = c(0.25, 0.5, 0.75), conf.int=FALSE)
```

All estimated values for survival function including point-wise confidence interval.

```{r}
summary(KM0)
```

### Plot estimated survival function

Global estimate including pointwise confidence intervals.

```{r rerSurvivalKM02}
plot(KM0, main=expression(paste("Kaplan-Meier-estimate ", hat(S)(t), " with CI")),
     xlab="t", ylab="Survival", lwd=2)
```

Separate estimates for levels of factor `IV`

```{r rerSurvivalKM03}
plot(KM, main=expression(paste("Kaplan-Meier-estimate ", hat(S)[g](t), " for groups g")),
     xlab="t", ylab="Survival", lwd=2, col=1:3)
legend(x="topright", col=1:3, lwd=2, legend=LETTERS[1:3])
```

### Plot cumulative hazard

\(\hat{\Lambda}(t)\)

```{r rerSurvivalKM04}
plot(KM0, main=expression(paste("Kaplan-Meier-estimate ", hat(Lambda)(t))),
     xlab="t", ylab="cumulative hazard", fun="cumhaz", lwd=2)
```

### Log-rank-test for equal survival-functions

Global test

```{r}
survdiff(Surv(obsT, status) ~ IV, data=dfSurv)
```

Stratified for factor `sex`

```{r}
survdiff(Surv(obsT, status) ~ IV + strata(sex), data=dfSurv)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:survival))
try(detach(package:splines))
```
