---
license: Creative Commons BY-SA
author: "Daniel Wollschlaeger"
title: "Survival analysis: Parametric proportional hazards models"
categories: [Univariate, Survival]
rerCat: Univariate
tags: [Survival]
---

Survival analysis: Parametric proportional hazards models
=========================

TODO
-------------------------

 - link to survivalKM, survivalCoxPH

Install required packages
-------------------------

[`survival`](https://cran.r-project.org/package=survival)

```{r}
wants <- c("survival")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Simulated right-censored event times with Weibull distribution
-------------------------

Simulated survival time \(T\) influenced by time independent covariates \(X_{j}\) with effect parameters \(\beta_{j}\) under assumption of proportional hazards, stratified by sex.

\(T = (-\ln(U) \, b \, e^{-\bf{X} \bf{\beta}})^{\frac{1}{a}}\), where \(U \sim \mathcal{U}(0, 1)\), \(a\) is the Weibull shape parameter and \(b\) is the Weibull scale parameter.

```{r}
set.seed(123)
N      <- 180                  # number of observations
P      <- 3                    # number of groups
sex    <- factor(sample(c("f", "m"), N, replace=TRUE))  # stratification factor
X      <- rnorm(N, 0, 1)       # continuous covariate
IV     <- factor(rep(LETTERS[1:P], each=N/P))  # factor covariate
IVeff  <- c(0, -1, 1.5)        # effects of factor levels (1 -> reference level)
Xbeta  <- 0.7*X + IVeff[unclass(IV)] + rnorm(N, 0, 2)
weibA  <- 1.5                  # Weibull shape parameter
weibB  <- 100                  # Weibull scale parameter
U      <- runif(N, 0, 1)       # uniformly distributed RV
eventT <- ceiling((-log(U)*weibB*exp(-Xbeta))^(1/weibA))   # simulated event time

# censoring due to study end after 120 days
obsLen <- 120                  # length of observation time
censT  <- rep(obsLen, N)       # censoring time = end of study
obsT   <- pmin(eventT, censT)  # observed censored event times
status <- eventT <= censT      # has event occured?
dfSurv <- data.frame(obsT, status, sex, X, IV)          # data frame
```

Parametric Proportional Hazards Models
-------------------------

Special cases of the Cox proportional hazards model that follow from assuming a specific form for the baseline hazard \(\lambda_{0}(t)\).

### Exponential distribution

\(E(T) = b > 0\), density function \(f(t) = \lambda(t) \, S(t)\) with constant hazard-function \(\lambda(t) = \lambda = \frac{1}{b}\) and survival-function \(S(t) = e^{-\frac{t}{b}}\). Cumulative hazard-function is \(\Lambda(t) = \frac{t}{b}\).

Alternative parametrization using rate parameter \(\lambda\):

\(f(t) = \lambda \, e^{-\lambda t}\), \(S(t) = e^{-\lambda t}\) and \(\Lambda(t) = \lambda t\). In functions `rexp()`, `dexp()` etc. option `rate` is \(\lambda\). The influence of covariates \(X_{j}\) causes \(\lambda' = \lambda \, e^{\bf{X} \bf{\beta}}\). 

Assumptions (note that \(\bf{X} \bf{\beta}\) does not include an intercept \(\beta_{0}\))

$$
\begin{equation*}
\begin{array}{rclcl}
\lambda(t)     &=& \frac{1}{b} \, e^{\bf{X} \bf{\beta}} &=& \lambda \, e^{\bf{X} \bf{\beta}}\\
\ln \lambda(t) &=& -\ln b + \bf{X} \bf{\beta} &=& \ln \lambda + \bf{X} \bf{\beta}\\
S(t)           &=& \exp\left(-\frac{t}{b} \, e^{\bf{X} \bf{\beta}}\right) &=& \exp\left(-\lambda \, t \, e^{\bf{X} \bf{\beta}}\right)\\
\Lambda(t)     &=& \frac{t}{b} \, e^{\bf{X} \bf{\beta}} &=& \lambda \, t \, e^{\bf{X} \bf{\beta}}
\end{array}
\end{equation*}
$$

### Weibull distribution

Parametrization used by `rweibull()`, `dweibull()` etc.:

Shape parameter \(a > 0\), scale parameter \(b > 0\), such that \(f(t) = \lambda(t) \, S(t)\) with hazard-function \(\lambda(t) = \frac{a}{b} \left(\frac{t}{b}\right)^{a-1}\) and survival-function \(S(t) = \exp(-(\frac{t}{b})^{a})\). Cumulative hazard-function is \(\Lambda(t) = (\frac{t}{b})^{a}\) with inverse \(\Lambda^{-1}(t) = (b \, t)^{\frac{1}{a}}\). \(E(T) = b \, \Gamma(1 + \frac{1}{a})\). The exponential distribution is a special case for \(a = 1\).

Alternative parametrization:

Setting \(\lambda = \frac{1}{b^{a}}\), such that \(\lambda(t) = \lambda \, a \, t^{a-1}\), \(S(t) = \exp(-\lambda \, t^{a})\) and \(\Lambda(t) = \lambda \, t^{a}\). The influence of covariates \(X_{j}\) causes \(\lambda' = \lambda \, e^{\bf{X} \bf{\beta}}\).

Assumptions (note that \(\bf{X} \bf{\beta}\) does not include an intercept \(\beta_{0}\))

$$
\begin{equation*}
\begin{array}{rclcl}
\lambda(t)     &=& \frac{a}{b} \left(\frac{t}{b}\right)^{a-1} \, e^{\bf{X} \bf{\beta}} &=& \lambda \, a \, t^{a-1} \, e^{\bf{X} \bf{\beta}}\\
\ln \lambda(t) &=& \ln \left(\frac{a}{b} \left(\frac{t}{b}\right)^{a-1}\right) + \bf{X} \bf{\beta} &=& \ln \lambda + \ln a + (a-1) \, \ln t + \bf{X} \bf{\beta}\\
S(t)           &=& \exp\left(-(\frac{t}{b})^{a} \, e^{\bf{X} \bf{\beta}}\right) &=& \exp\left(-\lambda \, t^{a} \, e^{\bf{X} \bf{\beta}}\right)\\
\Lambda(t)     &=& (\frac{t}{b})^{a} \, e^{\bf{X} \bf{\beta}} &=& \lambda \, t^{a} \, e^{\bf{X} \bf{\beta}}
\end{array}
\end{equation*}
$$

Parameter tests and model comparisons
-------------------------

### Parameter tests

Weibull-model

```{r}
library(survival)                     # for survreg()
fitWeib <- survreg(Surv(obsT, status) ~ X + IV, dist="weibull", data=dfSurv)
summary(fitWeib)
```

Transform AFT parameter \(\hat{\gamma}_{j} = - \hat{\beta}_{j} \cdot \hat{a}\) to parameter \(\hat{\beta}_{j}\) a Cox PH model.

```{r}
(betaHat <- -coef(fitWeib) / fitWeib$scale)
```

### Model comparisons

Exponential model = restricted Weibull model with shape parameter \(a = 1\).

```{r}
fitExp <- survreg(Surv(obsT, status) ~ X + IV, dist="exponential", data=dfSurv)
anova(fitExp, fitWeib)               # model comparison
```

Test significance of factor `IV` (associated with multiple effect parameters) as a whole by doing a model comparison against the restricted model without factor `IV`.

```{r}
# restricted model without IV
fitR <- survreg(Surv(obsT, status) ~ X, dist="weibull", data=dfSurv)
anova(fitR, fitWeib)                 # model comparison
```

Estimate survival-function
-------------------------

Apply fit to new data with specific observations. Result from `predict()` is the estimated distribution function \(\hat{F}^{-1}(t)\) for given percentiles `p`.

```{r}
dfNew <- data.frame(sex=factor(c("m", "m"), levels=levels(dfSurv$sex)),
                      X=c(0, 0),
                     IV=factor(c("A", "C"), levels=levels(dfSurv$IV)))
percs <- (1:99)/100
FWeib <- predict(fitWeib, newdata=dfNew, type="quantile", p=percs, se=TRUE)
```

To plot estimated survival function for two new observations, calculate \(S(t) = 1-F(t)\) and use `newdata` argument.

```{r rerSurvivalParametric01}
matplot(cbind(FWeib$fit[1, ],
              FWeib$fit[1, ] - 2*FWeib$se.fit[1, ],
              FWeib$fit[1, ] + 2*FWeib$se.fit[1, ]), 1-percs,
        type="l", main=expression(paste("Weibull-Fit ", hat(S)(t), " mit SE")),
        xlab="t", ylab="Survival", lty=c(1, 2, 2), lwd=2, col="blue")
matlines(cbind(FWeib$fit[2, ],
               FWeib$fit[2, ] - 2*FWeib$se.fit[2, ],
               FWeib$fit[2, ] + 2*FWeib$se.fit[2, ]), 1-percs, col="red", lwd=2)
legend(x="topright", lwd=2, lty=c(1, 2, 1, 2), col=c("blue", "blue", "red", "red"),
       legend=c("sex=m, X=0, IV=A", "+- 2*SE", "sex=m, X=0, IV=C", "+- 2*SE"))
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:survival))
```
