---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Survival analysis: Cox proportional hazards model"
categories: [Univariate, Survival]
rerCat: Univariate
tags: [Survival, CoxProportionalHazards]
---

Survival analysis: Cox proportional hazards model
=========================

TODO
-------------------------

 - link to survivalKM, survivalParametric, counting process notation for recurrent events and time dependent covariates 

Install required packages
-------------------------

[`survival`](http://cran.r-project.org/package=survival)

```{r}
wants <- c("survival")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Simulated right-censored event times with Weibull distribution
-------------------------

Time independent covariates with proportional hazards

\(T = (-\ln(U) \, b \, e^{-\bf{X} \bf{\beta}})^{\frac{1}{a}}\), where \(U \sim \mathcal{U}(0, 1)\)

```{r}
set.seed(1.234)
N      <- 180
P      <- 3
sex    <- factor(sample(c("f", "m"), N, replace=TRUE))
X      <- rnorm(N, 0, 1)
IV     <- factor(rep(LETTERS[1:P], each=N/P))
IVeff  <- c(0, -1, 1.5)
Xbeta  <- 0.7*X + IVeff[unclass(IV)] + rnorm(N, 0, 2)
weibA  <- 1.5
weibB  <- 100
U      <- runif(N, 0, 1)
eventT <- ceiling((-log(U)*weibB*exp(-Xbeta))^(1/weibA))
obsLen <- 120
censT  <- rep(obsLen, N)
obsT   <- pmin(eventT, censT)
status <- eventT <= censT
dfSurv <- data.frame(obsT, status, sex, X, IV)
```

Fit the Cox proportional hazards model
-------------------------

Assumptions

$$
\begin{equation*}
\begin{array}{rclcl}
\ln \lambda(t) &=& \ln \lambda_{0}(t) + \beta_{1} X_{1} + \dots + \beta_{p} X_{p}      &=& \ln \lambda_{0}(t) + \bf{X} \bf{\beta}\\
    \lambda(t) &=&     \lambda_{0}(t) \, e^{\beta_{1} X_{1} + \dots + \beta_{p} X_{p}} &=& \lambda_{0}(t) \, e^{\bf{X} \bf{\beta}}\\
S(t)           &=& S_{0}(t)^{\exp(\bf{X} \bf{\beta})} &=& \exp\left(-\Lambda_{0}(t) \, e^{\bf{X} \bf{\beta}}\right)\\
\Lambda(t)     &=& \Lambda_{0}(t) \, e^{\bf{X} \bf{\beta}} & &
\end{array}
\end{equation*}
$$

```{r}
library(survival)
(fitCPH <- coxph(Surv(obsT, status) ~ X + IV, data=dfSurv))
summary(fitCPH)
```

Assess model fit
-------------------------

### AIC

```{r}
library(survival)
extractAIC(fitCPH)
```

### McFadden, Cox & Snell and Nagelkerke pseudo \(R^{2}\)

Log-likelihoods for full model and 0-model without predictors X1, X2

```{r}
LLf <- fitCPH$loglik[2]
LL0 <- fitCPH$loglik[1]
```

McFadden pseudo-\(R^2\)

```{r}
as.vector(1 - (LLf / LL0))
```

Cox & Snell

```{r}
as.vector(1 - exp((2/N) * (LL0 - LLf)))
```

Nagelkerke

```{r}
as.vector((1 - exp((2/N) * (LL0 - LLf))) / (1 - exp(LL0)^(2/N)))
```

### Model comparisons

Restricted model without factor `IV`

```{r}
fitCPH1 <- coxph(Surv(obsT, status) ~ X, data=dfSurv)
anova(fitCPH1, fitCPH)          # model comparison
```

Estimate survival-function und cumulative baseline hazard
-------------------------

### Survival function

For average pseudo-observation with covariate values equal to the sample means. For factors: means of dummy-coded indicator variables.

```{r}
library(survival)                # for survfit()
(CPH <- survfit(fitCPH))

## survival 2.37-2 has a bug in quantile() so this currently doesn't work
# quantile(CPH, probs=c(0.25, 0.5, 0.75), conf.int=FALSE)
```

More meaningful: Estimated survival function for new specific data

```{r rerSurvivalCoxPH01}
dfNew  <- data.frame(sex=factor(c("f", "f"), levels=levels(dfSurv$sex)),
                       X=c(-2, -2),
                      IV=factor(c("A", "C"), levels=levels(dfSurv$IV)))
CPHnew <- survfit(fitCPH, newdata=dfNew)

par(mar=c(5, 4.5, 4, 2)+0.1, cex.lab=1.4, cex.main=1.4)
plot(CPH, main=expression(paste("Cox PH-estimate ", hat(S)(t), " with CI")),
     xlab="t", ylab="Survival", lwd=2)
lines(CPHnew$time, CPHnew$surv[ , 1], lwd=2, col="blue")
lines(CPHnew$time, CPHnew$surv[ , 2], lwd=2, col="red")
legend(x="topright", lwd=2, col=c("black", "blue", "red"),
       legend=c("pseudo-observation", "sex=f, X=-2, IV=A", "sex=f, X=-2, IV=C"))
```

### Cumulative baseline hazard

```{r rerSurvivalCoxPH02}
library(survival)                # for basehaz()
expCoef  <- exp(coef(fitCPH))
Lambda0A <- basehaz(fitCPH, centered=FALSE)
Lambda0B <- expCoef[2]*Lambda0A$hazard
Lambda0C <- expCoef[3]*Lambda0A$hazard
plot(hazard ~ time, main=expression(paste("Cox PH-estimate ", hat(Lambda)[g](t), " per group")),
     type="s", ylim=c(0, 5), xlab="t", ylab="cumulative hazard", lwd=2, data=Lambda0A)
lines(Lambda0A$time, Lambda0B, lwd=2, col="red")
lines(Lambda0A$time, Lambda0C, lwd=2, col="green")
legend(x="bottomright", lwd=2, col=1:3, legend=LETTERS[1:3])
```

Model diagnostics
-------------------------

### Proportional hazards assumption

Test based on scaled Schoenfeld-residuals

```{r}
library(survival)                      # for cox.zph()
(czph <- cox.zph(fitCPH))
```

Plot scaled Schoenfeld-residuals against covariates

```{r rerSurvivalCoxPH03}
par(mfrow=c(2, 2), cex.main=1.4, cex.lab=1.4)
plot(czph)
```

### Influential observations

```{r rerSurvivalCoxPH04}
dfbetas <- residuals(fitCPH, type="dfbetas")

par(mfrow=c(2, 2), cex.main=1.4, cex.lab=1.4)
plot(dfbetas[ , 1], type="h", main="DfBETAS for X",    ylab="DfBETAS", lwd=2)
plot(dfbetas[ , 2], type="h", main="DfBETAS for IV-B", ylab="DfBETAS", lwd=2)
plot(dfbetas[ , 3], type="h", main="DfBETAS for IV-C", ylab="DfBETAS", lwd=2)
```

### Linearity of log hazard

Plot martingale-residuals against continuous covariate with a nonparametric LOESS-smoother

```{r rerSurvivalCoxPH05}
resMart <- residuals(fitCPH, type="martingale")
plot(dfSurv$X, resMart, main="Martingale-residuals for X",
     xlab="X", ylab="Residuen", pch=20)
lines(loess.smooth(dfSurv$X, resMart), lwd=2, col="blue")
legend(x="bottomleft", col="blue", lwd=2, legend="LOESS fit", cex=1.4)
```

Predicted hazard ratios
-------------------------

With respect to an average pseudo-observation with covariate values equal to the sample means. For factors: means of dummy-coded indicator variables.

```{r}
library(survival)
hazRat <- predict(fitCPH, type="risk")
head(hazRat)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:survival))
try(detach(package:splines))
```
