---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Variance, robust spread measures, skewness and kurtosis"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

Variance, robust spread measures, skewness and kurtosis
=========================

TODO
-------------------------

 - link to diagDistributions, varianceHom

Install required packages
-------------------------

[`DescTools`](http://cran.r-project.org/package=DescTools), [`robustbase`](http://cran.r-project.org/package=robustbase)

```{r}
wants <- c("DescTools", "robustbase")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Variance and standard deviation
-------------------------

### Corrected (sample) variance and standard deviation

```{r}
age <- c(17, 30, 30, 25, 23, 21)
N   <- length(age)
M   <- mean(age)
var(age)
sd(age)
```

### Uncorrected (population) variance and standard deviation

```{r}
(cML <- cov.wt(as.matrix(age), method="ML"))
(vML <- diag(cML$cov))
sqrt(vML)
```

Robust spread measures
-------------------------

### Winsorized variance and standard deviation

```{r}
library(DescTools)
ageWins <- Winsorize(age, probs=c(0.2, 0.8))
var(ageWins)
sd(ageWins)
```

### Inter-quartile-range

```{r}
quantile(age)
IQR(age)
```
### Mean absolute difference to the median

```{r}
library(DescTools)
MeanAD(age)
```

### Median absolute difference to the median (MAD)

```{r}
mad(age)
```

### \(Q_{n}\): more efficient alternative to MAD

```{r}
library(robustbase)
Qn(age)
```

### \(\tau\) estimate of scale

```{r}
scaleTau2(age)
```

Diversity of categorical data
-------------------------

```{r}
fac <- factor(c("C", "D", "A", "D", "E", "D", "C", "E", "E", "B", "E"),
              levels=c(LETTERS[1:5], "Q"))
P   <- nlevels(fac)
(Fj <- prop.table(table(fac)))
```

First, calculate Shannon index, then diversity measure.

```{r}
library(DescTools)
shannonIdx <- Entropy(Fj, base=exp(1))
(H <- (1/log(P)) * shannonIdx)
```

Higher moments: skewness and kurtosis
-------------------------

```{r}
library(DescTools)
Skew(age)
Kurt(age)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:robustbase))
try(detach(package:DescTools))
```
