---
layout: post
title: "Variance, robust spread measures, skewness and kurtosis"
categories: [Descriptive]
rerCat: Descriptive
tags: [Descriptive]
---

Variance, robust spread measures, skewness and kurtosis
=========================

TODO
-------------------------

 - link to diagDistributions, varianceHom

Install required packages
-------------------------

[`e1071`](http://cran.r-project.org/package=e1071), [`psych`](http://cran.r-project.org/package=psych), [`robustbase`](http://cran.r-project.org/package=robustbase), [`vegan`](http://cran.r-project.org/package=vegan)

```{r}
wants <- c("e1071", "psych", "robustbase", "vegan")
has   <- wants %in% rownames(installed.packages())
if(any(!has)) install.packages(wants[!has])
```

Variance and standard deviation
-------------------------

### Corrected (sample) variance and standard deviation

```{r}
age <- c(17, 30, 30, 25, 23, 21)
N   <- length(age)
M   <- mean(age)
var(age)
sd(age)
```

### Uncorrected (population) variance and standard deviation

```{r}
(cML <- cov.wt(as.matrix(age), method="ML"))
(vML <- diag(cML$cov))
sqrt(vML)
```

Robust spread measures
-------------------------

### Winsorized variance and standard deviation

```{r}
library(psych)
ageWins <- winsor(age, trim=0.2)
var(ageWins)
sd(ageWins)
```

### Inter-quartile-range

```{r}
quantile(age)
IQR(age)
```
### Mean absolute difference to the median

```{r}
mean(abs(age-median(age)))
```

### Median absolute difference to the median (MAD)

```{r}
mad(age)
```

### \(Q_{n}\): more efficient alternative to MAD

```{r}
library(robustbase)
Qn(age)
```

### \(\tau\) estimate of scale

```{r}
scaleTau2(age)
```

Diversity of categorical data
-------------------------

```{r}
fac <- factor(c("C", "D", "A", "D", "E", "D", "C", "E", "E", "B", "E"),
              levels=c(LETTERS[1:5], "Q"))
P   <- nlevels(fac)
(Fj <- prop.table(table(fac)))
```

```{r}
library(vegan)
shannonIdx <- diversity(Fj, index="shannon")
(H <- (1/log(P)) * shannonIdx)
```

Higher moments: skewness and kurtosis
-------------------------

```{r}
library(e1071)
skewness(age)
kurtosis(age)
```

Detach (automatically) loaded packages (if possible)
-------------------------

```{r}
try(detach(package:psych))
try(detach(package:robustbase))
try(detach(package:vegan))
try(detach(package:permute))
try(detach(package:e1071))
try(detach(package:class))
```
