---
license: Creative Commons BY-SA
author: Daniel Wollschlaeger
title: "Exploring the workspace"
categories: [RBasics]
rerCat: R_Basics
tags: [Workspace, Environment]
---

Exploring the workspace
=========================

The working directory
-------------------------

```{r}
getwd()
```

```{r eval=FALSE}
setwd("d:/daniel/work/r/")
```

Create and remove objects
-------------------------

Create some objects

```{r}
Aval <- 7
Bval <- 15
Cval <- 10
ls()
```

List available objects

```{r}
# objects in .GlobalEnv workspace that have a capital C in their name
ls(".GlobalEnv", pattern="C")
exists("Bval")
exists("doesNotExist")
```

Remove objects

```{r}
Aval
rm(Aval)
Aval                             # this will give an error
## rm(list=ls(all.names=TRUE))   # remove all objects from the workspace
```

Show and rename objects
-------------------------

### Show objects

```{r}
Bval
print(Bval)
(Bval <- 4.5)
.Last.value
```

```{r}
get("Bval")
varName <- "Bval"
get(varName)
```

### Rename objects

```{r}
ls()
newNameVar <- "varNew"
assign(newNameVar, Bval)
varNew
```

Environments and search path
-------------------------

```{r}
search()
```

[How R Searches and Finds Stuff](http://obeautifulcode.com/R/How-R-Searches-And-Finds-Stuff/)

Session history
-------------------------

By default, `savehistory()` saves the command history to the file `.Rhistory` in the current working directory.

```{r eval=FALSE}
history()
savehistory("d:/daniel/work/r/history.r")
```

By default, `save.image()` saves all objects in the workspace to the file `.RData` in the current working directory.

```{r eval=FALSE}
save.image("d:/daniel/work/r/objects.Rdata")
```
